---
title: "Mapping INBO color ring data to SOVON format"
author: Damiano Oldoni
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
github-repo: inbo/cr-birding
output:
  bookdown::gitbook:
    split_by: rmd
    df_print: paged
    config:
      search: no
      sharing:
        facebook: no
        twitter: yes
        all: []
---

```{r index-setup, include = FALSE}
knitr::opts_chunk$set(
  warning = TRUE,
  message = TRUE,
  rownames.print = FALSE # Option for df_print = paged
)
```

# Intro

This document contains and describes all the steps in mapping INBO color ring data to SOVON format. By running this document in R, it will create all the necessary data files. Settings to access INBO database are necessary.

Load libraries:

```{r load_libraries}
library(DBI) # To connect to database
library(odbc) # To connect to database
library(tidyverse) # To do datascience
library(tidylog) # To provide feedback on dplyr functions
library(tidyselect) # To select columns smartly
library(measurements) # To convert measurement units
library(here) # To find files
library(lubridate) # To ease working with dates
```

The output of the data processing consists of three dataframes saved as comma separated files (`csv`) in folder `data/processed/`:

1. `crbirding_users`: user data (not loaded to GitHub for privacy reasons)
2. `crbirding_birds`: ring data
3. `crbirding_observations`: observation data
