---
title: "Import and preprocess INBO color ring data"
author: 
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("./docs/",sub(".Rmd", ".html", basename(input_file))))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

This pipeline will:

1. extract data from INBO *kleurring* database
2. perform some checks and data cleaning
3. save temporary data as text files

# Extract data from INBO kleurring database

## Connection to INBO database

We establish a connection to the database. Sensible data are saved in the `Renviron.site` file:

```{r connect_to_db}
conn <- dbConnect(odbc::odbc(),
                  driver = Sys.getenv("driver"),
                  server = Sys.getenv("server"),
                  database = Sys.getenv("database"),
                  port = Sys.getenv("port"),
                  trusted_connection = Sys.getenv("tc"))
```

## Import data

### Extract user data

First, we extract data about users from INBO *kleurring* database:

```{r get_user_data}
users <-  dbGetQuery(conn, "SELECT * FROM dbo.tblWaarnemer") %>%
    as_tibble()
```

### Extract color ring data

```{r get_birds_data}
birds <- dbGetQuery(conn, "SELECT * FROM dbo.tblKleurring") %>%
  as_tibble()
```

### Extract observation data

INBO's observation data contain a text type field: `Opmerking`. Text type is deprecated and an error will be returned if we perform the standard SQL query `"SELECT * FROM dbo.tblWaarneming"`. So, we need an ad-hoc query:

```{r get_obs_data}
obs <- dbGetQuery(conn,
  "SELECT Nummer,
          Datum,
          EuringCode,
          LeeftijdCode,
          KleurringNummer,
          KleurringPlaats,
          MetaalringNummer,
          MetaalringPlaats,
          PlaatsGemeente,
          PlaatsToponym,
          PlaatsToponymDetail,
          Convert(nvarchar(4000),Opmerking) as Opmerking,
          WaarnemerNummer,
          PlaatsLengtegraadGraden,
          PlaatsLengtegraadMinuten,
          PlaatsLengtegraadSeconden,
          PlaatsBreedtegraadGraden,
          PlaatsBreedtegraadMinuten,
          PlaatsBreedtegraadSeconden,
          PlaatsLengtegraadRichtingCode,
          PlaatsBreedtegraadRichtingCode,
          PlaatsLandCode,
          MetaalringLandCode,
          BevestigingDatum,
          PlaatsProvincie,
          AanmaakDatum,
          WijzigDatum
  FROM dbo.tblWaarneming") %>% as_tibble()
```

Table `tblWaarnemingAktie` is also important because contains informations about the actions taken at each observation:

```{r get_waarnemingAktie}
obs_actions <- dbGetQuery(conn, "SELECT * FROM dbo.tblWaarnemingAktie")
```

The action codes are described in table `dbo.cdeAktie`:

```{r get_}
actions_meaning <- dbGetQuery(conn, "SELECT * FROM dbo.cdeAktie")
actions_meaning
```

Close connection to server:

```{r close_connection}
dbDisconnect(conn)
```

### Import control data

From UVABIT repository, we copied a text file containing the mapping of GPS ids:

```{r import_GPS_names_ids}
map_gps_path <- here::here("data", "external", "map_gps_id_to_color_ring.tsv")
map_gps <- 
  read_tsv(map_gps_path) %>%
  mutate(bird_notes = "uvabits_gps_tag.")
map_gps %>% head()
```

This file will be used later in this same document.

# Pre-processing

## Preview raw data

For privcay reasons users data cannot be shown. Users data refer to the following fields:

```{r prview_users}
colnames(users)
```

Preview birds data:

```{r preview_birds}
birds %>% head()
```

Preview observations data:

```{r preview_obs}
obs %>% head()
```

Preview observation actions data:

```{r prview_obs_actions}
obs_actions %>% head()
```

## Check spaces in `birds`

No spaces should be present in any column of `birds`:

```{r check_spaces}
map_dfr(birds, ~ mutate(birds, space_detect = str_detect(., pattern = " "))) %>%
  filter(space_detect == TRUE)
```

Remove all spaces:

```{r remove_spaces}
birds <- birds %>% 
  mutate_all(funs(str_replace(., " ", "")))
```

## Check duplicates `WaarnemingNummer`-`AktieCode`

The action acronym is contained in column `AktieCode`. Observations and their actions are linked via columns `WaarnemingNummer` (in `actions`) and `Nummer` (in `obs`).

No duplicates `WaarnemingNummer`-`AktieCode` should exist:

```{r show_duplicates_actions}
obs_actions %>% 
  filter(WaarnemingNummer %in% 
           (obs_actions %>% 
              group_by(WaarnemingNummer, AktieCode) %>% 
              count() %>%
              filter(n > 1) %>%
              pull(WaarnemingNummer))
)
```

Otherwise they should be removed:

```{r to_be_removed}
obs_actions <- 
  obs_actions %>% 
  distinct(WaarnemingNummer, AktieCode, .keep_all = TRUE)
```

## Merge observations and actions

In order to ease the mapping of birds and observations, we first remove columns will be never used: 

```{r remove_cols_not_used}
obs_actions <- 
  obs_actions %>% 
  select(-c(Nummer,
            AanmaakDatum, 
            WijzigDatum)
)
```

Then we *spread* action codes to columns:

```{r }
obs_actions <- 
  obs_actions %>%
  spread(AktieCode, AktieCode)
```

The following combinations occurr:

```{r distinct_combinations_actions}
combinations_actions <- obs_actions %>%
  select(-WaarnemingNummer) %>%
  distinct()
combinations_actions
```

The following actions are at least once present in database:

```{r colnames_actions_present}
acts <- names(combinations_actions)
```

Actions described in `actions_meaning` never used:

```{r no_used_actions}
actions_meaning %>%
  filter(!Code %in% acts)
```

We add action code columns to observations in order to have a unique dataframe:

```{r merge_obs_with_acts}
obs_and_acts <- 
  obs %>% 
  left_join(obs_actions, by = c("Nummer" = "WaarnemingNummer"))
```

Preview: 

```{r merge_obs_with_acts}
head(obs_and_acts, n = 100)
```

## Check spaces in `obs_and_acts`

No spaces should be present in character columns of `obs_and_acts` except for columns related to place description (`Plaats*`), datums (`*Datum`) and notes (`Opmerking`):

```{r show_spaces_obs_and_acts}
map_dfr(obs_and_acts %>%
          select_if(is.character) %>%
          select(-c(starts_with("Plaats"), 
                    Opmerking, 
                    ends_with("Datum"))), ~ 
          mutate(obs_and_acts, space_detect = str_detect(., pattern = " "))) %>%
  filter(space_detect == TRUE) %>%
  arrange(Nummer)
```

Remove spaces:

```{r remove_spaces_from_obs_and_acts}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate_at(vars(-c(starts_with("Plaats"),
                 ends_with("Datum"),
                 one_of("Opmerking", "Nummer", "WaarnemerNummer"))), 
            funs(str_replace(., " ", ""))
)
```

## Check presence action code

All observations should also have an action code:

```{r observations_without_action}
no_actions <- 
  obs_and_acts %>%
  filter_at(vars_select(names(obs_and_acts), one_of(acts)), all_vars(is.na(.)))
no_actions
```

Observation without it will be removed:

```{r remove_observations_without_action}
obs_and_acts <- 
  obs_and_acts %>% 
  filter(!Nummer %in% no_actions$Nummer)
```

## Remove data labelled as `FOUT`

There are observations judged as error. They are marked by`KleurringNummmer` equal to `FOUT`:

```{r show_FOUT_obs}
obs_and_acts %>% 
  filter(KleurringNummer == "FOUT") %>%
  head(n = 10)
```

The value `FOUT` is also present in `birds`:

```{r show_FOUT_birds}
birds %>% 
  filter(Nummer == "FOUT")
```

These data are related to observations judged not correct (`FOUT`: _error_) by the INBO experts and administrator of the database.

Thus, at the moment, we remove these data (discussion about this issue [here](https://github.com/inbo/sovon/issues/9)) from observations:

```{r remove_FOUT_obs}
obs_and_acts <- 
  obs_and_acts %>% 
  filter(KleurringNummer != "FOUT")
```

and from birds: 

```{r remove_FOUT_birds}
birds <- 
  birds %>% filter(Nummer != "FOUT")
```

## Check color rings: uppercase

Bird color rings codes are uppercase. Exceptions:

```{r show_rings_lowercase}
birds %>%
  filter(Nummer != toupper(Nummer) | 
           NummerNieuw != toupper(NummerNieuw))
```

are converted to uppercase:

```{r convert_rings_to_uppercase}
birds <- birds %>%
  mutate(Nummer = toupper(Nummer),
         NummerNieuw = toupper(NummerNieuw),
         NummerDesc = toupper(NummerDesc))
```

The same holds true for `KleurringNummer` in `obs_and_acts`. Exceptions:

```{r lowercase_kleurringNummer}
obs_and_acts %>% 
  filter(KleurringNummer != toupper(KleurringNummer)) %>% 
  distinct(KleurringNummer)
```

are transformed to uppercase:

```{r KleurringNummer_to_upper}
obs_and_acts <- obs_and_acts %>% 
  mutate(KleurringNummer = toupper(KleurringNummer))
```

## Check keys `KleurringNummer` (obs) - `Nummer` (birds)

All values in `KleurringNummer` should be present in `birds$Nummer`. Exceptions:

```{r KleurringNummer_not_in_birds_Nummer}
obs_and_acts %>%
  filter(!KleurringNummer %in% birds$Nummer) %>% 
  distinct(KleurringNummer)
```

We remove them:

```{r remove_obs_and_acts_linked_to_invalid_bird_ring}
KleurringNummer_remove <- 
  obs_and_acts %>%
  filter(!KleurringNummer %in% birds$Nummer) %>% 
  distinct(KleurringNummer) %>%
  pull()
obs_and_acts <- 
  obs_and_acts %>% 
  filter(!KleurringNummer %in% KleurringNummer_remove)
```

Finally, we search for birds (`Nummer` in  `birds`) not linked to any observation (`KleurringNummer` in `obs_and_acts`):

```{r birds_not_in_obs}
birds %>% filter(!Nummer %in% obs_and_acts$KleurringNummer)
```

We remove them:

```{r remove_birds_not_in_obs}
birds <- 
  birds %>% 
  filter(Nummer %in% obs_and_acts$KleurringNummer)
```

## Check duplicates in date of applying rings

Detect duplicates in date of applying (the very first) color ring (action code `rngkl`):

```{r duplicates_apply_first_color_ring}
duplicates_ringing_rngkl <- 
  obs_and_acts %>%
  filter(!is.na(rngkl)) %>%
  distinct(KleurringNummer, Datum, rngkl)  %>%
  group_by(KleurringNummer, Datum, rngkl) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(obs_and_acts,
            by = c("KleurringNummer", "Datum", "rngkl")) %>%
  select(Nummer, everything())
duplicates_ringing_rngkl
```

and metal ring (action code `rngme`):

```{r duplicates_apply_first_metal_ring}
duplicates_ringing_rngme <- 
  obs_and_acts %>%
  filter(!is.na(rngkl)) %>%
  distinct(KleurringNummer, Datum, rngme)  %>%
  group_by(KleurringNummer, Datum, rngme) %>%
  count() %>%
  filter(n > 1) %>%
  select(-n) %>%
  left_join(obs_and_acts,
            by = c("KleurringNummer", "Datum", "rngme")) %>%
  select(Nummer, everything())
duplicates_ringing_rngme
```

## Check age while applying rings

For each bird, observations with same date should refer to same bird age. Exceptions:

```{r check_anomalies_age_ringing}
obs_and_acts %>%
  filter(!is.na(LeeftijdCode)) %>%
  filter(!is.na(rngkl) | !is.na(rngme)) %>%
  distinct(KleurringNummer, Datum, LeeftijdCode)  %>%
  group_by(KleurringNummer, Datum) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(obs_and_acts %>%
              filter(!is.na(LeeftijdCode)) %>%
              select(Nummer, KleurringNummer, Datum, 
                     LeeftijdCode, rngkl, rngme)) %>%
  select(-n) %>%
  select(Nummer, KleurringNummer, Datum, 
         LeeftijdCode, rngkl, rngme
)
```

## Check bird sex inconsistencies

All birds should have a sex, which could be one of `M` (_mannetje_), `V` (_vrouwtje_), `O` (_onbekend_):

```{r present_GeslachtCode}
distinct(birds, GeslachtCode)
```

Exceptions:

```{r exceptions_bird_sex}
birds %>%
  filter(is.na(GeslachtCode)) %>%
  distinct(Nummer, NummerNieuw, NummerDesc, GeslachtCode)
```

For all birds a one-to-one relation bird - sex should hold true:

```{r check_one-to-one_bird_sex}
birds %>% 
  distinct(Nummer, GeslachtCode) %>%
  nrow() == nrow(birds)
```

## Handle information about metal ring number

There are columns containing metal ring related information in `birds`:

```{r cols_metal_ring_birds}
names(birds)[which(str_detect(names(birds), "Metaalring"))]
```

and `obs_and_acts`:

```{r cols_metal_ring_obs_and_acts}
names(obs_and_acts)[which(str_detect(names(obs_and_acts), "Metaalring"))]
```

This is not only redundant: inconsistencies are detected. Some examples:

```{r inconsisntecy_metal_ring_info_examples}
birds %>%
  filter(Nummer %in% c("AAAA", "AAAR", "AAAT", "AAAW")) %>%
  rename("metal_ring_from_birds"  = "MetaalringNummer") %>%
  left_join(obs_and_acts %>%
              rename("metal_ring_from_obs"  = "MetaalringNummer"),
            by = c("Nummer" =  "KleurringNummer")) %>%
  distinct(Nummer, metal_ring_from_birds, metal_ring_from_obs)
```

Based on experts' knowledge only the metal ring information in `birds` is correct. We rename the columns related to metal ring information from `obs_and_acts` by adding suffix `_obs`:

```{r remove_metal_ring_infos_from_obs}
obs_and_acts <-
  obs_and_acts %>%
  rename_at(vars(starts_with("Metaalring")),
            ~ paste0(., "_obs"))
```

## Solve GPS tracker IDs

Some birds have a suspect `Nummer` which is formed by letters `GPS` or `GP` followed by some numbers. These are a kind of GPS IDs which should not be in columns related to color rings:

```{r show_gps_ids}
birds %>%
  filter(str_detect(Nummer, pattern = "(^(GP)\\d+)|(^(GPS)\\d+)")) %>%
  select(Nummer)
```

The right mapping of them is saved in `map_gps`:

```{r preview_map_gps}
map_gps %>% head()
```

We import the color rings in new column `first_Nummer`. For all other rings will be `first_nummer = Nummer` as they are the very first rings:

```{r map_to_first_Nummer}
birds <- 
  birds %>%
  # create column first_Nummer with right color rings for the GPS or GP** risngs
  left_join(map_gps, by = c("Nummer" = "gps_id")) %>%
  # set first_Numer equal to Nummer in all other cases
  mutate(first_Nummer = ifelse(is.na(color_ring),
                               Nummer,
                               color_ring)) %>%
  select(Nummer, first_Nummer, everything())
```

Check whether not corrected rings (`GP***` or `GPS***`) are still present:

```{r check_presence_GP_GPS_in_birds}
birds %>%
  filter(str_detect(first_Nummer, pattern = "(^(GP)\\d+)|(^(GPS)\\d+)")) %>%
  select(first_Nummer)
```

Mapping example: the bird with `Nummer` equal to `GP130` is mapped as follows:

```{r example_gps_mapping_and_name}
birds %>%
  filter(Nummer == "GP130") %>%
  select(Nummer, first_Nummer, bird_notes)
```

Column `first_Nummer` should contain unique elements. Check for exceptions:

```{r check_uniqueness_first_Nummer}
birds %>%
  group_by(first_Nummer) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(birds) %>%
  select(-n)
```

Exceptions are due to double entry, one with the original color ring, a second one with GPS tracker ID. Such exceptions, if present, should be handled separately while mapping `bird_shorthand` in document `3_birds.Rmd`.

## Check inconsistencies between users and observers

We have to be sure that all observations have a valid observator reference, i.e. an ID contained in column `Nummer` of `users`.

Observations without an observator reference:

```{r WaarnemerNummer_is_na}
obs_and_acts %>%
  filter(is.na(WaarnemerNummer))
```

All observator references in `obs_and_acts` should be present in `users`:

```{r check_observators_in_users}
observers_ids <- unique(
  obs_and_acts %>%
    filter(!is.na(WaarnemerNummer)) %>%
    distinct(WaarnemerNummer) %>%
    pull())
all(observers_ids %in% users$Nummer)
```

# Save preprocessed data

After data cleaning we save the data as TSVs in `.data/interim`:

```{r}
write_tsv(users, path = here::here("data", "interim", "users.tsv"), na = "")
write_tsv(birds, path = here::here("data", "interim", "birds.tsv"), na = "")
write_tsv(
  obs_and_acts, 
  path = here::here("data", "interim", "obs_and_actions.tsv"), 
  na = "")
```

We save the table containing actions code and relative meaning in `./data/input` as it could be useful in following mapping steps:

```{r save_actions_meaning}
write_tsv(
  actions_meaning, 
  path = here::here("data", "input", "actions_meaning.tsv"), 
  na = ""
)
```

