# Map observation data

This pipeline will map observation data.

# Read data

## Read temporary observation data

We start from the tempoary observation data saved as TSV file in `data\interim`:

```{r import_temp_observations}
obs_and_acts <- read_tsv(
  here::here("data", "interim", "obs_and_actions.tsv"), 
  guess_max = 500000
)
```

## Read action data

Import action codes and relative meaning:

```{r import_actions_meaning}
actions_meaning <- read_tsv(here::here("data", "input", "actions_meaning.tsv"))
```

## Read processed ring data

We import finalized ring data from `./data/processed`:

```{r import_final_ring_data}
crbirding_birds <- read_tsv(
  here::here("data", "processed", "crbirding_birds.tsv"), 
  guess_max = 50000
)
```

## Read temporary ring data

We will need some columns from original INBO ring data. We import the temporary ring data as well:

```{r import_birds_from_INBO_temp_data}
birds <- read_tsv(
  here::here("data", "interim", "birds.tsv"), 
  guess_max = 50000
)
```

## Read processed user data

We import finalized user data from `./data/processed`:

```{r import_final_ring_data}
crbirding_users <- read_tsv(
  here::here("data", "processed", "crbirding_users.tsv"), 
  guess_max = 50000
)
```

## Read list workers in bird shelters

```{r import_workers}
workers_bird_shelters <- read_tsv(
  here::here("data", "input", "workers_in_bird_shelters.tsv")
)
```

# Map color observation data

## Extract action codes

Actions present in `obs_and_acts` :

```{r}
acts <- actions_meaning$Code
acts <- acts[acts %in% names(obs_and_acts)]
acts
```

## Observation reference

The observation reference is an unique identifier assigned to each observation. This field exists already: `Nummer`.

```{r add_obs_ref}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(sovon_observation_reference = Nummer)
```

## Observation date

The date is saved in column `Datum`. We copy it in required column `sovon_observation_date`:

```{r obs_date}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(sovon_observation_date = Datum)
```

## Observation time

There is no observation time in `obs_and_acts`. `NA` is given to `sovon_observation_time`:

```{r obs_time}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(sovon_observation_time = NA)
```

## Observation latitude

Observation latitude should be converted to decimal degrees:

```{r obs_latitude}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(
    sovon_observation_lat = conv_unit(
      str_c(PlaatsBreedtegraadGraden,
            PlaatsBreedtegraadMinuten,
            PlaatsBreedtegraadSeconden, 
            sep = " "), 
      from = 'deg_min_sec', to = 'dec_deg')
)
```

## Observation longitude

Observation longitude should be converted to decimal degrees:

```{r obs_longitude}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(
    sovon_observation_long = conv_unit(
      str_c(PlaatsLengtegraadGraden,
            PlaatsLengtegraadMinuten,
            PlaatsLengtegraadSeconden, 
            sep = " "), 
      from = 'deg_min_sec', to = 'dec_deg')
)
```

## Observation location

Aggregate information about observation location. We follow the following structure: `PlaatsGemeente` [+ `, ` + `PlaatsToponym` [+ `: ` + `PlaatsToponymDetail`]]:

```{r obs_location}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(
    sovon_observation_location = ifelse(!is.na(PlaatsToponym),
                                  str_c(PlaatsGemeente,
                                        PlaatsToponym, 
                                        sep = ", "),
                                  paste(PlaatsGemeente))) %>%
  mutate(
    sovon_observation_location = ifelse(!is.na(PlaatsToponymDetail),
                                  str_c(sovon_observation_location,
                                        PlaatsToponymDetail,
                                        sep = ": "),
                                  paste(sovon_observation_location)))
```

## Check bird

The field `sovon_check_bird` is provided by SOVON. `NA` is given:

```{r check_bird}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_check_bird = NA)
```

## Observation reporter reference number

The field `sovon_melder_ringersnummer` is equal to `user_reference`:

```{r sovon_melder_ringersnummer}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_melder_ringersnummer = WaarnemerNummer)
```

## Observation reporter

The field `sovon_melder` is equal to `user_first_name` and `user_last_name` in `users`:

```{r obs_sovon_melder}
obs_and_acts <- obs_and_acts %>%
  left_join(
    crbirding_users %>% 
      select(user_reference, user_first_name, user_last_name),
    by = c("WaarnemerNummer" = "user_reference")) %>%
  mutate(sovon_melder = str_c(user_first_name, user_last_name, sep = " ")) %>%
  select(-c(user_first_name, user_last_name))
```

## Ring number

The field `sovon_ring_number` contains information about the metal ring number, in the same format as in field `bird_ring_number` of `crbirding_birds`. As explained in the mapping of field `bird_ring_number` of `crbirding_birds`, INBO database is a color ring database. This field is not accurately mapped and we will leave it empty.

```{r add_ring_number}
obs_and_acts <-
  obs_and_acts %>%
  mutate(sovon_ring_number = NA_character_)
```

## Observation status

Adding the EURING status information arises by the need of mapping the observations with action code `br` and `vang`:

```{r breed_show}
actions_meaning %>%
  filter(Code %in% c("br", "vang"))
```

The actions `br` and `vang` refer to status `breeding` in EURING system. Based on the scheme at page 16 of [EURING Exchange Code 2000+](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) document, we define the following mapping:

```{r obs_status}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_status = case_when(
    !is.na(br) | !is.na(vang) ~ "N",
    LeeftijdCode == "PU" ~ "-",
    TRUE ~ "U"
  ))
```

## Condition

The SOVON field `condition` is mapped following the scheme at page 22 of [EURING Exchange Code 2000+](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) document. The following actions are mapped as follows:
1. `dood`: condition = 1
2. `vang`: condition = 8
3. `vangl`: condition = 8
4. `veld`: condition = 7
5. `me`: condition = 7
6. `ziek`: condition = 4, 5

```{r obs_condition}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_condition = case_when(
    !is.na(dood) ~ 1,
    !is.na(vang) | !is.na(vangl) ~ 8,
    !is.na(veld) ~ 7,
    !is.na(me) ~ 7,
    !is.na(ziek) ~ 4)
)
```

The observations of sick birds not anymore released should get condition equal to 5. These observations are linked to the very last `ziek` action and followed by an observation with action `dood`. Not only, but the observer should be a worker of a bird shelter, in `workers_bird_shelters`.

Reference number of observations with condition 5:

```{r get_obs_cond_5}
last_sick_df <- 
  obs_and_acts %>%
  filter(ziek == "ziek") %>%
  group_by(sovon_bird_reference) %>%
  filter(sovon_observation_date == max(sovon_observation_date)) %>%
  ungroup() %>%
  select(sovon_observation_reference, sovon_bird_reference, sovon_observation_date) %>%
  rename(last_sick_date = sovon_observation_date,
         obs_ref = sovon_observation_reference)
dead_df <-
  obs_and_acts %>%
  filter(dood == "dood") %>%
  select(sovon_bird_reference, sovon_observation_date) %>%
  rename(dead_date = sovon_observation_date)
last_sick_df <- 
  last_sick_df %>%
  filter(sovon_bird_reference %in% dead_df$sovon_bird_reference) %>%
  left_join(dead_df, by = "sovon_bird_reference")

condition_5_obs_ref <- 
  obs_and_acts %>%
  filter(sovon_bird_reference %in% last_sick_df$sovon_bird_reference) %>%
  left_join(last_sick_df,
            by = "sovon_bird_reference") %>%
  group_by(sovon_bird_reference) %>%
  filter(sovon_observation_date >= last_sick_date & 
           sovon_observation_date < dead_date) %>%
  count() %>%
  # only one ziek action present
  filter(n == 1) %>%
  left_join(last_sick_df,
            by = c("sovon_bird_reference")) %>%
  pull(obs_ref)
```

Assign condition value `5` to these observations:

```{r assign_5}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_condition = 
           if_else(
             sovon_observation_reference %in% condition_5_obs_ref, 
             5,
             sovon_condition)
)
```

Preview:

```{r preview}
obs_and_acts %>%
  filter(sovon_condition == 5) %>%
  select(sovon_observation_reference, 
         sovon_bird_reference, 
         sovon_observation_date,
         acts)
```

## MRI: metal ring information

Metal ring information is an integer between 0 and 9. A table can be found at page 8 of [EURING Exchange Code 2000+](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) document.

However, SOVON uses this field to map the observations referring to the applying of any kind of ring, either color or metal, as their database is purely focussed on color rings and due to a missing field in the EURING Exchange Code 2000+ specific for color rings.

We initialize `sovon_MRI` as an empty column:

```{r intialize_MRI}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = NA_integer_)
```

Position of metal ring:

```{r}
birds %>% 
  distinct(MetaalringPlaats)
```

where `LL` means *left leg* and  `RL` means *right leg*. So, we exclude MRI values 2, *Metal ring added, definitely on tarsus*, and 3, *Metal ring added, definitely above tarsus*.

All observations linked to `rngme` or `rngkl` get MRI equal to 1:

```{r set_MRI_1}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(!is.na(rngme) | !is.na(rngkl), 1, sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_1}
obs_and_acts %>%
  filter(sovon_MRI == 1) %>%
  select(sovon_observation_reference, sovon_MRI, rngme, rngkl) %>%
  head(n = 50)
```

MRI is 4 for observations linked to action `me` (metal ring read)  `veld` (field observation, equiavelnt of colro ring read):

```{r MRI_4_me_veld}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(!is.na(me) | !is.na(veld), 4, sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_4}
obs_and_acts %>%
  filter(sovon_MRI == 4) %>%
  select(sovon_observation_reference, sovon_MRI, me, veld) %>%
  head(n = 50)
```

MRI is 0 for observations of birds without any ring, thus linked to actions `meweg` (metal ring missing) and `klweg` (color ring missing):

```{r MRI_0_meweg}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(!is.na(meweg) & !is.na(klweg), 0 , sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_0}
obs_and_acts %>%
  filter(sovon_MRI == 0) %>%
  select(sovon_observation_reference, sovon_MRI, me, veld) %>%
  head(n = 50)
```

The actions `vang` related to a change of color ring can be found by checking the datum of the observation joined with table `crbirdin_birds`:

```{r vang_with_change_color_ring}
obs_and_acts %>%
  mutate(sovon_MRI = if_else(!is.na(vang))
```


By screening the texual notes of observations related to `vang` with 6 or more consecutive numbers, we can see which observations refer to changing the metal ring (MRI equal to 5), which ones refer to apply a new metal ring as the old one is not present anymore (MRI equal to 1), and which ones confirm the presence of metal ring without any change (MRI equal to 4):

```{r vang_notes}
vang_notes <- 
  obs_and_acts %>%
  filter(str_detect(Opmerking, pattern = "[0-9]{6,}")) %>%
  filter(AktieCode == "vang") %>%
  select(sovon_observation_reference, Opmerking, bird_reference, Datum) %>%
  arrange(bird_reference)
vang_notes
```

Observations linked to `vang` where a metal ring is added (old metal ring absent):

```{r vang_metal_ring_added}
vang_notes_metalring_added <- 
  vang_notes %>%
  filter(Opmerking %in%
           c("AFAB > LBAW + metaal L907345 (oud metaal was niet meer aanwezig)",
             "oud metaal L89066 was niet aanwezig, L913152 aangebracht",	
             "ASAP --> SZAT en H902297",
             "oud metaal niet meer aanwezig? (H106600) > H901760\r\nASAV > PMAV",
             "L89137 verloren --> L914601, AVAD -> UTAL",
             "L89303 was reeds verloren > L909956, BBAX -> KZAC",
             "BHAW --> URAA en L914607, bloedstaal genomen, oude metaal ring L89377 verloren",
             "gevangen op nest EA676412  toegvoegd",
             "oude metaalring L89460 verloren, nieuwe metaalring L909611 aangelegd",
             "L89160 was niet meer aanwezig ---> nieuw metaal is L914621",
             "BSAY ---> TSAK en L912663",
             "CFAC -> YNAJ, kleurring bijna doorgeslsten. NB oude metaalring verloren nieuw=H901953",
             "eerste metaalring verloren, wordt L904816"
             )
)
```

where the metal ring is present and not changed:

```{r vang_metal_ring_not_chagend}
vang_notes_metalring_present_no_changed <- 
  vang_notes %>%
  filter(Opmerking %in% 
           c("we B[DPAS] met L.905880 op het nest teruggevangen (vleugel 423, geslacht vrouw) en deze heeft nu B[HAAB] gekregen (staal niet vervangen)."))
```

and all the others referring to `vang` actions where metal ring has been changed:

```{r vang_metal_ring_changed}
vang_notes_metalring_changed <- 
  anti_join(vang_notes, 
            bind_rows(vang_notes_metalring_added, vang_notes_metalring_present_no_changed),
            by = names(vang_notes)
)
```

Check we covered all cases:

```{r check_nrows_vang_notes}
nrow(vang_notes) == 
  (nrow(vang_notes_metalring_added) + 
     nrow(vang_notes_metalring_present_no_changed) +
     nrow(vang_notes_metalring_changed))
```

Assign MRI equal to 1 to observations linked to `vang` where a metal ring is added (old metal ring absent):

```{r MRI_1_vang}
obs_ref_vang_ring_added <- 
  vang_notes_metalring_added %>%
  pull(sovon_observation_reference)

obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(sovon_observation_reference %in% obs_ref_vang_ring_added, 1, sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_1_vang}
obs_and_acts %>%
  filter(sovon_observation_reference %in% obs_ref_vang_ring_added) %>%
  select(sovon_observation_reference, bird_reference, AktieCode, sovon_MRI) %>%
  head(n = 50)
```

Assign MRI equal to 4 to observations linked to `vang` where a metal ring is already present and not changed:

```{r MRI_4_vang}
obs_ref_vang_ring_present <- 
  vang_notes_metalring_present_no_changed %>%
  pull(sovon_observation_reference)

obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(sovon_observation_reference %in% obs_ref_vang_ring_present, 4, sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_4_vang}
obs_and_acts %>%
  filter(sovon_observation_reference %in% obs_ref_vang_ring_present) %>%
  select(sovon_observation_reference, bird_reference, AktieCode, sovon_MRI)
```

Assign MRI equal to 5 to observations linked to `vang` where a metal ring is changed:

```{r MRI_5_vang}
obs_ref_vang_ring_changed <- 
  vang_notes_metalring_changed %>%
  pull(sovon_observation_reference)

obs_and_acts <- 
  obs_and_acts %>%
  mutate(sovon_MRI = ifelse(sovon_observation_reference %in% obs_ref_vang_ring_changed, 5, sovon_MRI))
```

Preview of changes:

```{r preview_changes_MRI_5_vang}
obs_and_acts %>%
  filter(sovon_observation_reference %in% obs_ref_vang_ring_changed) %>%
  select(sovon_observation_reference, bird_reference, AktieCode, sovon_MRI)
```

The MRI chronology:

```{r}
MRI_chronology <- 
  obs_and_acts %>% 
  filter(!is.na(sovon_MRI)) %>%
  distinct(bird_reference, sovon_observation_reference, sovon_MRI, sovon_observation_date) %>%
  arrange(bird_reference, sovon_observation_date)
MRI_chronology
```

We correct MRI when a 5 follows a 0: it should be 1. This is the case of observations linked to `vang` actions occurring just after `meweg` actions.

```{r correct_5_to_1}
MRI_chronology <-
  MRI_chronology %>%
  mutate(sovon_MRI = ifelse(sovon_MRI == 5 & lag(sovon_MRI) == 0,
                       sovon_MRI == 1, sovon_MRI))
```

Now we can spread the table to create intervals for each row:

```{r spread_chronology}
MRI_chronology_spread <- 
  MRI_chronology %>%
  group_by(bird_reference) %>%
  mutate(MRI_after = lead(sovon_MRI),
         MRI_after_sovon_observation_date = lead(sovon_observation_date)) %>%
  rename(MRI_before = sovon_MRI,
         MRI_before_sovon_observation_date = sovon_observation_date,
         observation_chronology = sovon_observation_reference)
MRI_chronology_spread
```

We can now assess the MRI value of all observations with missing MRI for each bird based on these rules:

1. observation taken after MRI 1 or 5 and before MRI 0: set MRI = 4
2. observation taken after MRI 0 and before MRI 1: set MRI = 0

```{r assign_last_MRI}
# obs_and_acts_assign_last_MRI <- 
obs_and_acts_test <-
  obs_and_acts %>%
  left_join(MRI_chronology_spread,
            by = c("bird_reference")) %>%
  group_by(bird_reference) %>%
  mutate(sovon_MRI = case_when(
    sovon_observation_date > MRI_before_sovon_observation_date & 
      (MRI_before == 1 | MRI_before == 5) & 
      ((MRI_after == 0 & sovon_observation_date < MRI_after_sovon_observation_date) | 
         is.na(MRI_after)) ~ 4,
    sovon_observation_date > MRI_before_sovon_observation_date & 
      MRI_before == 0 & 
      ((MRI_after == 1 & sovon_observation_date < MRI_after_sovon_observation_date) | 
         is.na(MRI_after)) ~ 0,
    TRUE ~ sovon_MRI)) %>%
  select(names(obs_and_acts))
duplicated(obs_and_acts_test) | duplicated(obs_and_acts_test, fromLast = TRUE)
nrow(obs_and_acts) == nrow(obs_and_acts_test)
```

## Observation notes

We copy the notes in `Opmerking` to SOVON field  `observation_notes`:

```{r obs_notes}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(observation_notes = Opmerking)
```

As other notes could follow, we add a dot at the end of the notes if not present already:

```{r}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(
    observation_notes = 
      if_else(
        !is.na(observation_notes) & str_length(observation_notes) > 0,
        if_else(str_sub(observation_notes, -1) != ".",
                str_c(observation_notes, ".", sep = ""),
                observation_notes),
        observation_notes)
)
```

We also add the prefix `"INBO original notes:"`:

```{r add_prefix}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(
    observation_notes = 
      if_else(
        !is.na(observation_notes) & str_length(observation_notes) > 0,
        str_c("INBO original notes:", observation_notes, sep = " "),
        observation_notes)
)
```

We have to add the note `no_color_ring.` to observations of birds not wearing any color ring (only metal ring), a situation tackled in previous section:

```{r add_note_no_color_ring}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(
    observation_notes = if_else(
      ,
      if_else(is.na(observation_notes),
              "no_color_ring.",
              paste(observation_notes, "no_color_ring.", sep = " ")),
      observation_notes)
)
```

Effects of mapping:

```{r}
obs_and_acts %>%
  filter(is.na(sovon_ring_number)) %>%
  select(sovon_observation_reference, sovon_ring_number, observation_notes)
```

# Observation data

Select the required columns, starting with prefix `sovon_`:

```{r select_sovoin_cols}
crbirding_observations <-
  crbirding_observations %>%
  select(starts_with("sovon"))
```

The desired order of columns in `crbirding_observations`:

```{r cols_order_obs}
cr_obs_cols <- c(
  "user_id", "user_reference", "bird_id", "bird_reference", "observation_id",
  "observation_reference", "observation_date", "observation_time", 
  "observation_lat", "observation_lng", "observation_location", 
  "observation_is_capture", "observation_notes", "check_bird", "MRI", "melder",
  "melder_email", "melder_ringersnummer", "ring_number", "condition", "status", 
  "bird_age_ringing", "bird_sex"
)
```

Are all required columns present?

```{r check_presence_required_cols_obs}
all(cr_obs_cols %in% names(crbirding_observations)) & 
  ncol(cr_obs_cols) == ncol(crbirding_observations)
```

Set column order:

```{r get_right_order_cols_obs}
crbirding_observations <-
  crbirding_observations %>%
  select(cr_obs_cols)
```

Preview data:

```{r final_preview_obs}
crbirding_observations %>% head(n = 10)
```

Save to CSV:

```{r write_processed_obs_data}
write_csv(
  crbirding_obs, 
  path = here::here("data", "processed", "crbirding_observations.tsv"), 
  na = ""
)
```
