# Compare to data transfer of 2019 and preserve IDs

SOVON asks us to preserve IDs used in mapped data as sent in April 2019.

## Read data

First, we need to read the data we sent them in April 2019.

```{r read_2019_data}
crbirding_users_2019 <- read_csv(here::here("data",
                                  "processed",
                                  "crbirding_users_20190410.csv"),
                                 col_types = cols(
                                   .default = col_character(),
                                   user_id = col_logical(),
                                   user_reference = col_number()),
                       na = "")
crbirding_birds_2019 <- read_csv(here::here("data",
                                  "processed",
                                  "crbirding_birds_20190410.csv"),
                       col_types = cols(
                         .default = col_character(),
                         bird_id = col_logical(),
                         bird_reference = col_number(),
                         bird_bto = col_logical(),
                         bird_birth_year = col_logical(),
                         bird_date_begin = col_datetime(
                           format = ""),
                         bird_date_end = col_datetime(
                           format = "")),
                       na = "")
crbirding_observations_2019 <- read_csv(here::here(
  "data",
  "processed",
  "crbirding_observations_20190410.csv"),
  col_types = cols(
    .default = col_character(),
    user_id = col_logical(),
    user_reference = col_number(),
    bird_id = col_logical(),
    bird_reference = col_number(),
    observation_id = col_logical(),
    observation_reference = col_number(),
    observation_date = col_datetime(
      format = ""),
    observation_time = col_logical(),
    check_bird = col_logical(),
    MRI = col_number(),
    melder_ringersnummer = col_logical(),
    condition = col_number()
  ),
  na = "")
```

We read also the data we mapped just now.

```{r read_data_now}
crbirding_users <- read_csv(here::here("data",
                                       "processed",
                                       "crbirding_users.csv"),
                            col_types = cols(
                              .default = col_character(),
                              user_id = col_logical(),
                              user_reference = col_number()),
                            na = "")
crbirding_birds <- read_csv(here::here("data",
                                       "processed",
                                       "crbirding_birds.csv"),
                            col_types = cols(
                              .default = col_character(),
                              bird_id = col_logical(),
                              bird_reference = col_number(),
                              bird_bto = col_logical(),
                              bird_birth_year = col_logical(),
                              bird_date_begin = col_datetime(
                                format = ""),
                              bird_date_end = col_datetime(
                                format = "")),
                            na = "")
crbirding_observations <- read_csv(here::here(
  "data",
  "processed",
  "crbirding_observations.csv"),
  col_types = cols(
    .default = col_character(),
    user_id = col_logical(),
    user_reference = col_number(),
    bird_id = col_logical(),
    bird_reference = col_number(),
    observation_id = col_logical(),
    observation_reference = col_number(),
    observation_date = col_datetime(
      format = ""),
    observation_time = col_logical(),
    check_bird = col_logical(),
    MRI = col_number(),
    melder_ringersnummer = col_logical(),
    condition = col_number()
  ),
  na = "")
```

## Users

Number of new users based on new `user_reference` (email and other private info not shown):

```{r new_user_reference}
crbirding_users %>%
  filter(!user_reference %in% crbirding_users_2019$user_reference) %>%
  select(user_reference, user_country, user_role)
```

Old `user_reference` IDs not anymore present (email and other private info not shown):

```{r old_user_ref_not_present}
crbirding_users_2019 %>%
  filter(!user_reference %in% crbirding_users$user_reference) %>%
  select(user_reference, user_country, user_role)
```

Users from 2019 where some changes have been applied:

```{r users_with_changes}
same_ref_ids <- 
  crbirding_users %>%
  filter(user_reference %in% crbirding_users_2019$user_reference) %>%
  pull(user_reference)
crbirding_users_2019 %>%
  filter(user_reference %in% same_ref_ids) %>%
  anti_join(crbirding_users,
            by = names(crbirding_users)) %>%
  pull(user_reference)
```

## Map birds

### Map `bird_reference`

In INBO ring table, birds were uniquely identified by their very first ring (data were spread and untidy). We added a `bird_reference` field to gather information about multiple rings and link them to the same bird. This means that this field cannot be used to compare the actual INBO data with the data of 2019. We have to compare data by the very first color ring.

```{r first_color_ring}
crbirding_birds_first_color_ring <-
  crbirding_birds %>% 
  select(bird_reference, bird_shorthand, bird_date_begin) %>%
  group_by(bird_reference) %>%
  filter(!is.na(bird_shorthand)) %>%
  mutate(first_color_ring = if_else(bird_date_begin == min(bird_date_begin,
                                                     na.rm = TRUE),
                              bird_shorthand,
                              NA_character_)) %>%
  ungroup() %>%
  filter(!is.na(first_color_ring)) %>%
  distinct(bird_reference, first_color_ring) %>%
  right_join(crbirding_birds, by = c("bird_reference"))
```

All birds should get a valid  `first_color_ring` column. Exceptions:

```{r birds_without_first_color_ring}
crbirding_birds_first_color_ring %>%
  filter(is.na(first_color_ring))
```

We do the same with ring data from 2019:

```{r first_color_ring_2019}
crbirding_birds_first_color_ring_2019 <-
  crbirding_birds_2019 %>% 
  select(bird_reference, bird_shorthand, bird_date_begin) %>%
  group_by(bird_reference) %>%
  filter(!is.na(bird_shorthand)) %>%
  mutate(first_color_ring = if_else(bird_date_begin == min(bird_date_begin,
                                                           na.rm = TRUE),
                                    bird_shorthand,
                                    NA_character_)) %>%
  ungroup() %>%
  filter(!is.na(first_color_ring)) %>%
  distinct(bird_reference, first_color_ring) %>%
  right_join(crbirding_birds_2019, by = c("bird_reference"))
```

Even in this case all birds should get a valid  `first_color_ring` column. Exceptions:

```{r birds_without_first_color_ring_2019}
crbirding_birds_first_color_ring_2019 %>%
  filter(is.na(first_color_ring))
```

Now we can map the new `bird_reference` to the `bird_reference` of 2019:

```{r mapping_bird_reference_new_old}
crbirding_birds_mapping_bird_reference <-
  crbirding_birds_first_color_ring %>%
  distinct(bird_reference, first_color_ring) %>%
  left_join(crbirding_birds_first_color_ring_2019 %>%
              distinct(bird_reference, first_color_ring) %>%
              rename(bird_reference_2019 = bird_reference),
            by = "first_color_ring") %>%
  select(first_color_ring, everything())
head(crbirding_birds_mapping_bird_reference, n = 100)
```

We add  `bird_reference_2019` to  `crbirding_birds`:

```{r add_bird_ref_2019}
crbirding_birds <- 
  crbirding_birds %>%
  left_join((crbirding_birds_mapping_bird_reference %>%
              select(bird_reference, bird_reference_2019)),
            by = "bird_reference") %>%
  select(bird_id, bird_reference, bird_reference_2019, everything())
```

Number of birds present in 2019 but with `bird_reference` not equal to `bird_reference_2019` :

```{r n_birds_changed_bird_reference}
crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  filter(bird_reference != bird_reference_2019) %>%
  nrow
```

This is not surprising as the bird_reference has always been defined as a progressive number, so it is sufficient the change of one bird reference to modify all the other references coming after.

First change appears at:

```{r first_change}
crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  filter(bird_reference != bird_reference_2019) %>%
  filter(bird_reference == min(bird_reference)) %>%
  select(bird_reference, bird_reference_2019)
```

### New `bird_reference`

Some birds where not present in data transfer of April 2019, so the `bird_reference` cannot be mapped to `bird_reference_2019`.

This case includes:

1. New birds: birds ringed for the very first time after the previous data transfer in April 2019.
2. brids with modified first color ring: birds whose very first color ring has been corrected. This case is due to the fact that the first color ring is the real unique identifier in INBO ring table `tblKleurring`

```{r new_birds_or_modified_first_ring}
crbirding_birds %>%
  # filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference"))
```

### Changed  `bird_euring` 

```{r changed_bird_euring}
crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_euring")) %>%
  select(bird_reference, bird_reference_2019, bird_euring) %>%
  left_join(crbirding_birds_2019 %>%
              rename(bird_euring_2019 = bird_euring) %>%
              select(bird_reference, bird_euring_2019),
            by = (c("bird_reference_2019" = "bird_reference")))
```

### Changed  `bird_shorthand` 

Birds with changed ring history:

```{r changed_bird_shorthand}
bird_ref_changed_shorthand <-
  crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_shorthand")) %>%
  distinct(bird_reference, bird_reference_2019)
bird_ref_changed_shorthand
```

`bird_shorthand` history in  `crbirding_birds`:

```{r different_bird_shorthand}
crbirding_birds %>%
  filter(bird_reference %in% bird_ref_changed_shorthand$bird_reference) %>%
  select(bird_reference,
         bird_reference_2019,
         bird_shorthand,
         bird_date_begin,
         bird_date_end)
```

Values of `bird_shorthand` in  `crbirding_birds_2019`:

```{r bird_shorthand_in_2019}
crbirding_birds_2019 %>%
  filter(bird_reference %in% bird_ref_changed_shorthand$bird_reference_2019) %>%
  select(bird_reference,
         bird_shorthand,
         bird_date_begin,
         bird_date_end) %>%
  rename_all(paste0, "_2019")
```

### Changed `bird_scheme`

Rings with changed `bird_scheme`:

```{r changed_bird_scheme}
changed_bird_scheme <- 
  crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_scheme")) %>%
  select(bird_reference, bird_reference_2019, bird_scheme, bird_date_begin, bird_date_end) %>%
  left_join(crbirding_birds_2019 %>%
              rename(bird_scheme_2019 = bird_scheme) %>%
              select(bird_reference,
                     bird_scheme_2019,
                     bird_date_begin,
                     bird_date_end),
            by = (c("bird_reference_2019" = "bird_reference",
                    "bird_date_begin", "bird_date_end"))) %>%
  filter(!is.na(bird_scheme_2019) | !is.na(bird_scheme))
changed_bird_scheme
```

Mapping values:

```{r mapping_changes_bird_scheme}
changed_bird_scheme %>%
  group_by(bird_scheme, bird_scheme_2019) %>%
  count()
```

### Changed `bird_ring_number`

Rings with changed `bird_ring_number`:

```{r changed_bird_ring_number}
changed_bird_ring_number <- 
  crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_ring_number")) %>%
  select(bird_reference, bird_reference_2019, bird_ring_number, bird_date_begin, bird_date_end) %>%
  left_join(crbirding_birds_2019 %>%
              rename(bird_ring_number_2019 = bird_ring_number) %>%
              select(bird_reference,
                     bird_ring_number_2019,
                     bird_date_begin,
                     bird_date_end),
            by = (c("bird_reference_2019" = "bird_reference",
                    "bird_date_begin", "bird_date_end"))) %>%
  filter(!is.na(bird_ring_number_2019) | !is.na(bird_ring_number)) %>%
  select(contains("reference"),
         contains("number"),
         contains("date"))
changed_bird_ring_number
```

### Changed `bird_name`

Rings with changed `bird_name`:

```{r changed_bird_name}
changed_bird_name <- 
  crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_name")) %>%
  select(bird_reference, 
         bird_reference_2019, 
         bird_name, 
         bird_date_begin,
         bird_date_end) %>%
  left_join(crbirding_birds_2019 %>%
              rename(bird_name_2019 = bird_name) %>%
              select(bird_reference,
                     bird_name_2019,
                     bird_date_begin,
                     bird_date_end),
            by = (c("bird_reference_2019" = "bird_reference",
                    "bird_date_begin", "bird_date_end"))) %>%
  filter(!is.na(bird_name_2019) | !is.na(bird_name)) %>%
  select(contains("reference"),
         contains("name"),
         contains("date"))
changed_bird_name
```

### Changed `bird_sex`

Rings with changed `bird_sex`:

```{r changed_bird_sex}
changed_bird_sex <- 
  crbirding_birds %>%
  filter(!is.na(bird_reference_2019)) %>%
  anti_join(crbirding_birds_2019,
            by = c("bird_reference_2019" = "bird_reference",
                   "bird_sex")) %>%
  select(bird_reference, 
         bird_reference_2019, 
         bird_sex, 
         bird_date_begin,
         bird_date_end) %>%
  left_join(crbirding_birds_2019 %>%
              rename(bird_sex_2019 = bird_sex) %>%
              select(bird_reference,
                     bird_sex_2019,
                     bird_date_begin,
                     bird_date_end),
            by = (c("bird_reference_2019" = "bird_reference",
                    "bird_date_begin", "bird_date_end"))) %>%
  filter(!is.na(bird_sex_2019) | !is.na(bird_sex)) %>%
  select(contains("reference"),
         contains("sex"),
         contains("date"))
changed_bird_sex
```

### Changed `bird_date_begin`

Rings with changed `bird_date_begin`:

```{r changed_bird_date_begin}
same_rings <- 
  crbirding_birds %>%
  select(bird_reference,
         bird_reference_2019,
         bird_shorthand,
         bird_date_begin,
         bird_date_end) %>%
  inner_join(crbirding_birds_2019 %>%
               select(bird_reference,
                      bird_date_begin_2019 = bird_date_begin,
                      bird_shorthand,
                      bird_date_end_2019 = bird_date_end),
             by = c("bird_reference_2019" = "bird_reference",
                    "bird_shorthand"))
changed_bird_date_begin_end <-
  same_rings %>%
  group_by(bird_reference) %>%
  filter(!bird_date_begin %in% bird_date_begin_2019 | 
           !bird_date_end %in% bird_date_end_2019) %>%
  filter(!is.na(bird_date_begin) &
           !is.na(bird_date_begin_2019) &
           !is.na(bird_date_end) &
           !is.na(bird_date_end_2019))
changed_bird_date_begin_end
```

### Save crbirding_birds with added reference from 2019 export

We save crbirding_birds with the added `bird_reference_2019`:

```{r save_birds_with_bird_ref_2019}
# transform dates to characters
crbirding_birds %>%
  mutate(
    bird_date_begin = as.Date(bird_date_begin),
    bird_date_end = as.Date(bird_date_end)) %>%
  write_csv(path = here::here("data",
                              "processed",
                              "crbirding_birds.csv"),
  na = ""
)
```

## Map observations

Number of new observations based on new `observation_reference` (email and other private info not shown):

```{r new_observation_reference}
crbirding_observations %>%
  filter(!observation_reference %in% 
           crbirding_observations_2019$observation_reference) %>%
  nrow()
```

Old observations not anymore present (all columns info from 2019 data):

```{r old_obs_ref_not_present}
crbirding_observations_2019 %>%
  filter(!observation_reference %in% 
           crbirding_observations$observation_reference) %>%
  select(-contains("melder"))
```

Save in `same_ref_ids` the observation reference of observations present in both data.frames:

```{r same_ref_ids_obs}
same_ref_ids <- 
  crbirding_observations %>%
  filter(observation_reference %in% crbirding_observations_2019$observation_reference) %>%
  pull(observation_reference)
```

### Changed geospatial information

Details of observations with changes in latitude (`observation_lat`) and/or longitude (`observation_lng`):

```{r obs_with_changes_lat_lng}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "observation_lat",
                   "observation_lng")) %>%
  select(observation_reference, observation_lat, observation_lng) %>%
  rename(observation_lat_2019 = observation_lat,
         observation_lng_2019 = observation_lng) %>%
  left_join(crbirding_observations %>%
              select(observation_reference,
                     observation_lat,
                     observation_lng),
            by = "observation_reference") %>%
  select(observation_reference,
         observation_lat_2019,
         observation_lat,
         observation_lng_2019,
         observation_lng)
```

Details of observations with changes in location (column `observation_location`):

```{r obs_with_changes_location}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "observation_location")) %>%
  select(observation_reference, observation_location) %>%
  rename(observation_location_2019 = observation_location) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, observation_location),
            by = "observation_reference")
```

### Changed temporal information

Details of observations with changes in temporal information (column `observation_date`):

```{r obs_with_changes_date}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "observation_date")) %>%
  select(observation_reference, observation_date) %>%
  rename(observation_date_2019 = observation_date) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, observation_date),
            by = "observation_reference")
```

### Changed user information

Details of observations with changes in user detail(no details about users shown, just `user_reference`):

```{r obs_with_changes_user}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "user_reference",
                   "melder",
                   "melder_email")) %>%
  select(observation_reference, user_reference) %>%
  rename(user_reference_2019 = user_reference) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, user_reference),
            by = "observation_reference")
```

### Changed metal ring information

Details of observations with changes in metal ring information (columns `MRI` and  `ring_number`):

```{r obs_with_changes_metalring_infos}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "MRI",
                   "ring_number")) %>%
  select(observation_reference, MRI, ring_number) %>%
  rename(MRI_2019 = MRI,
         ring_number_2019 = ring_number) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, MRI, ring_number),
            by = "observation_reference")
```

### Changed condition

Details of observations with changes in condition (column `condition`):

```{r obs_with_changes_condition}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "condition")) %>%
  select(observation_reference, condition) %>%
  rename(condition_2019 = condition) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, condition),
            by = "observation_reference")
```

### Changed status

First we rename column `status` of  `crbirding_observation_2019` to new column name `status_full_grown_bird`:

```{r rename_status}
crbirding_observations_2019 <-
  crbirding_observations_2019 %>%
  rename(status_full_grown_bird = status)
```

Details of observations with changes in status (column `status_full_grown_bird`):

```{r obs_with_changes_status}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "status_full_grown_bird")) %>%
  select(observation_reference, status_full_grown_bird) %>%
  rename(status_full_grown_bird_2019 = status_full_grown_bird) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, status_full_grown_bird),
            by = "observation_reference")
```

### Changed bird age

Details of observations with changes in bird age at time of observation (column `bird_age_obs`):

```{r obs_with_changes_bird_age_obs}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "bird_age_obs")) %>%
  select(observation_reference, bird_age_obs) %>%
  rename(bird_age_obs_2019 = bird_age_obs) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, bird_age_obs),
            by = "observation_reference")

```

### Changed bird sex

Details of observations with changes in bird sex at time of observation (column `bird_sex`):

```{r obs_with_changes_bird_sex}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "bird_sex")) %>%
  select(observation_reference, bird_sex) %>%
  rename(bird_sex_2019 = bird_sex) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, bird_sex),
            by = "observation_reference")

```

### Changed bird ring position

Details of observations with changes in bird ring position (column `bird_ring_position`) or direction (`bird_ring_direction`):

```{r obs_with_changes_bird_ring_position_direction}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference", 
                   "bird_ring_position",
                   "bird_ring_direction")) %>%
  select(observation_reference,
         bird_ring_position, 
         bird_ring_direction) %>%
  rename(bird_ring_position_2019 = bird_ring_position,
         bird_ring_direction_2019 = bird_ring_direction) %>%
  left_join(crbirding_observations %>%
              select(observation_reference,
                     bird_ring_position,
                     bird_ring_direction),
            by = "observation_reference") %>%
  select(observation_reference,
         bird_ring_position_2019,
         bird_ring_position,
         bird_ring_direction_2019,
         bird_ring_direction)
```

### Changed notes

Details of observations with changes in observation notes (column `observation_notes`):

```{r obs_with_changes_observation_notes}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "observation_notes")) %>%
  select(observation_reference, observation_notes) %>%
  rename(observation_notes_2019 = observation_notes) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, observation_notes),
            by = "observation_reference")
```

### Changed capture information

Details of observations with changes in capture information (column `observation_is_capture`):

```{r obs_with_changes_observation_is_capture}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("observation_reference",
                   "observation_is_capture")) %>%
  select(observation_reference, observation_is_capture) %>%
  rename(observation_is_capture_2019 = observation_is_capture) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, observation_is_capture),
            by = "observation_reference")
```

### Changed bird reference

Details of observations with changes in bird reference (column `bird_reference`):

```{r obs_with_changes_bird_reference}
crbirding_observations_2019 %>%
  filter(observation_reference %in% same_ref_ids) %>%
  anti_join(crbirding_observations,
            by = c("bird_reference",
                   "observation_reference")) %>%
  select(observation_reference, bird_reference) %>%
  rename(bird_reference_2019 = bird_reference) %>%
  left_join(crbirding_observations %>%
              select(observation_reference, bird_reference),
            by = "observation_reference")
```

The huge number of changes of `bird_reference` has been discussed in previous session related to `crbirding_birds`.
