---
title: "Map data from INBO color ring database to crbirding (SOVON)"
author: 
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("./docs/",sub(".Rmd", ".html", basename(input_file))))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

This pipeline will:

1. extract data from INBO *kleurring* database (official name: `D0016_00_Meeuwen`)
2. transform data to SOVON format
3. export data in text files

# Setup

Load libraries:

```{r load_libraries}
library(DBI) # To connect to database
library(odbc) # To connect to database
library(stringr) # To perform string operations
library(stringi) # To perform string operations
library(readr) # To read and write txt files
library(dplyr) # To transform data
library(tidyr) # To transform data
library(purrr) # To transform data via loops
library(measurements) # To convert units of measurement 
library(here) # To find files
library(lubridate) # To ease working with dates
```

Set output file paths:

```{r file_paths}
cr_users_file <- here::here("data", "processed", "data_users.tsv")
cr_birds_file <- here::here("data", "processed", "data_birds.tsv")
cr_obs_file <- here::here("data", "processed", "data_observations.tsv")
```

Set column orders of output files:

```{r cols_order}
# users
cr_users_cols <- c(
  "user_id", "user_reference", "user_email", "user_first_name",
  "user_last_name", "user_address", "user_postal_code", "user_place",
  "user_country", "user_language", "user_role"
)
# birds
cr_birds_cols <- c(
  "bird_id", "bird_reference", "bird_euring", "bird_bto", 
  "bird_shorthand", "bird_scheme", "bird_ring_number", "bird_name", "bird_sex",
  "bird_birth_year", "bird_date_begin", "bird_date_end", "bird_age_ringing"
)
# observations
cr_obs_cols <- c(
  "user_id", "user_reference", "bird_id", "bird_reference", "observation_id",
  "observation_reference", "observation_date", "observation_time", 
  "observation_lat", "observation_lng", "observation_location", 
  "observation_is_capture", "observation_notes", "check_bird", "MRI", "Melder",
  "Melder_email", "Melder_ringersnummer", "Ringnumber", "Conditie", 
  "Bird_age_ringing", "Bird_sex"
)
```

# Extract data from INBO kleurring database

## Connection to INBO database

We establish a connection to the database. Sensible data are saved in the `.Renviron` file:

```{r connect_to_db}
conn <- dbConnect(odbc::odbc(),
                  driver = Sys.getenv("driver"),
                  server = Sys.getenv("server"),
                  database = Sys.getenv("database"),
                  port = Sys.getenv("port"),
                  trusted_connection = Sys.getenv("tc"))
```

## Extract data

### Extract user data

First, we extract data about users from INBO *kleurring* database:

```{r get_user_data}
users <-  dbGetQuery(conn, "SELECT * FROM dbo.tblWaarnemer") %>%
    as_tibble()
```

### Extract color ring data

```{r get_birds_data}
birds <- dbGetQuery(conn, "SELECT * FROM dbo.tblKleurring") %>%
  as_tibble()
```

### Extract observation data

INBO's observation data contain a text type field: `Opmerking`. Text type is deprecated and an error will be returned if we perform the standard SQL query `"SELECT * FROM dbo.tblWaarneming"`. So, we need an ad-hoc query:

```{r get_obs_data}
obs <- dbGetQuery(conn,
  "SELECT Nummer,
          Datum,
          EuringCode,
          LeeftijdCode,
          KleurringNummer,
          KleurringPlaats,
          MetaalringNummer,
          MetaalringPlaats,
          PlaatsGemeente,
          PlaatsToponym,
          PlaatsToponymDetail,
          Convert(nvarchar(4000),Opmerking) as Opmerking,
          WaarnemerNummer,
          PlaatsLengtegraadGraden,
          PlaatsLengtegraadMinuten,
          PlaatsLengtegraadSeconden,
          PlaatsBreedtegraadGraden,
          PlaatsBreedtegraadMinuten,
          PlaatsBreedtegraadSeconden,
          PlaatsLengtegraadRichtingCode,
          PlaatsBreedtegraadRichtingCode,
          PlaatsLandCode,
          MetaalringLandCode,
          BevestigingDatum,
          PlaatsProvincie,
          AanmaakDatum,
          WijzigDatum
  FROM dbo.tblWaarneming") %>% as_tibble()
```

Table `tblWaarnemingAktie` is also important because contains informations about the actions the observations refer to:

```{r get_waarnemingAktie}
obs_actions <- dbGetQuery(conn, "SELECT * FROM dbo.tblWaarnemingAktie")
```

The action codes are described in table `dbo.cdeAktie`:

```{r get_}
actions_meaning <- dbGetQuery(conn, "SELECT * FROM dbo.cdeAktie")
actions_meaning
```

### Import control data

The INBO seebird team provided a spreadsheet to map color rings (see section [bird shorthand](#bird-shorthand)). From this spreadhseet we distilled some text files with specific mapping information:

```{r import_csv_from_control_spreadsheet}
map_gps_path <- here::here("data", "input", "map_gps_id_to_color_ring.tsv")
map_gps <- read_tsv(map_gps_path)
map_gps %>% head()
```

## Pre-processing

### Add prefix `raw_` to columns

Add prefix `raw_` to all column names to avoid any possible name clash with SOVON terms (for more info see [spreadsheet](https://docs.google.com/spreadsheets/d/16928bSQzZ0jz0FCf5gXZ4Rg4gf9qYQPO5izLZ5IYSTQ/edit?usp=drive_web&ouid=109004011205942623101) and [document](https://docs.google.com/document/d/1VtwGD4wrwJLnGnvJprtFB3MbbXdirMELYvuCGEhLr3w/edit)) as provided by SOVON (access restricted).

```{r add_raw}
colnames(users) <- paste0("raw_", colnames(users))
colnames(birds) <- paste0("raw_", colnames(birds))
colnames(obs) <- paste0("raw_", colnames(obs))
colnames(obs_actions) <- paste0("raw_", colnames(obs_actions))
```

### Preview data

For privcay reasons users data cannot be shown. Users data refer to the following fields:

```{r prview_users}
colnames(users)
```

Preview birds data:

```{r preview_birds}
birds %>% head()
```

Preview observations data:

```{r preview_obs}
obs %>% head()
```

Preview observation actions data:

```{r prview_obs_actions}
obs_actions %>% head()
```

### Check duplicates `raw_WaarnemingNummer`-`raw_AktieCode`

The action acronym is contained in column `raw_AktieCode`. Observations and their action are linked via columns `raw_WaarnemingNummer` (in `obs_actions`) and `raw_Nummer` (in `obs`).

No duplicates `raw_WaarnemingNummer`-`raw_AktieCode` should exist:

```{r show_duplicates_obs_actions}
obs_actions %>% filter(raw_WaarnemingNummer %in% 
                         (obs_actions %>% 
                         group_by(raw_WaarnemingNummer, 
                                  raw_AktieCode) %>% 
                         count() %>%
                         filter(n > 1) %>%
                         pull(raw_WaarnemingNummer)))
```

Otherwise they should be removed:

```{r to_be_removed_after}
obs_actions <- obs_actions %>% 
  distinct(raw_WaarnemingNummer, 
           raw_AktieCode, .keep_all = TRUE)
```

### Check presence action code

All observations should also have an action code:

```{r observations_without_action}
obs %>% 
  filter(!raw_Nummer %in% 
           (obs_actions %>% pull(raw_WaarnemingNummer)))
```

Observation without it will be removed:

```{r remove_observations_without_action}
obs <- obs %>% 
  filter(raw_Nummer %in% 
           (obs_actions %>% pull(raw_WaarnemingNummer)))
```

### Check spaces in `birds`

No spaces should be present in any column of `birds`:

```{r check_spaces}
map_dfr(birds, ~ mutate(birds, space_detect = str_detect(., pattern = " "))) %>%
  filter(space_detect == TRUE)
```

Remove all spaces:

```{r remove_spaces}
birds <- birds %>% 
  mutate_all(funs(str_replace(., " ", "")))
```

### Merge observations and actions

In order to ease the mapping of birds and observations, we add action codes to observations in order to have a unique **tidy** data frame:

```{r merge_obs_with_acts}
obs_and_acts <- obs %>% 
  left_join(obs_actions %>% 
              select(-c(raw_Nummer, 
                        raw_AanmaakDatum, 
                        raw_WijzigDatum)),
            by = c("raw_Nummer" = "raw_WaarnemingNummer"))
```

### Check spaces in `obs_and_acts`

No spaces should be present in character columns of `obs_and_acts` except for columns related to place description (`raw_Plaats*`), datums (`*Datum`) and notes (`raw_Opmerking`):

```{r show_spaces_obs_and_acts}
map_dfr(obs_and_acts %>%
          select_if(is.character) %>%
          select(-c(starts_with("raw_Plaats"), 
                    raw_Opmerking, 
                    ends_with("Datum"))), ~ 
          mutate(obs_and_acts, space_detect = str_detect(., pattern = " "))) %>%
  filter(space_detect == TRUE) %>%
  arrange(raw_Nummer)
```

Remove spaces:

```{r remove_spaces_from_obs_and_acts}
obs_and_acts <- obs_and_acts %>% 
  mutate_at(vars(-c(starts_with("raw_Plaats"),
                 ends_with("Datum"),
                 one_of("raw_Opmerking", "raw_Nummer", "raw_WaarnemerNummer"))), 
            funs(str_replace(., " ", "")))
```

### Remove data labelled as `FOUT`

There are observations judged as error. They are marked by`raw_KleurringNummmer` equal to `FOUT`:

```{r show_FOUT_obs}
obs_and_acts %>% 
  filter(raw_KleurringNummer == "FOUT") %>%
  head(n = 10)
```

The value `FOUT` is also present in `birds`:

```{r show_FOUT_birds}
birds %>% 
  filter(raw_Nummer == "FOUT")
```

These data are related to observations judged not correct (`FOUT`: _error_) by the INBO experts and administrator of the database.

Thus, at the moment, we remove these data (discussion about this issue [here](https://github.com/inbo/sovon/issues/9)):

```{r remove_FOUT}
obs_and_acts <- obs_and_acts %>% filter(raw_KleurringNummer != "FOUT")
birds <- birds %>% filter(raw_Nummer != "FOUT")
```

### Check color rings: uppercase

Bird color rings codes are uppercase. Exceptions:

```{r show_rings_lowercase}
birds %>%
  filter(raw_Nummer != toupper(raw_Nummer) | 
           raw_NummerNieuw != toupper(raw_NummerNieuw))
```

are converted to uppercase:

```{r convert_rings_to_uppercase}
birds <- birds %>%
  mutate(raw_Nummer = toupper(raw_Nummer),
         raw_NummerNieuw = toupper(raw_NummerNieuw),
         raw_NummerDesc = toupper(raw_NummerDesc))
```

The same holds true for `raw_KleurringNummer` in `obs_and_acts`. Exceptions:

```{r lowercase_raw_kleurringNummer}
obs_and_acts %>% 
  filter(raw_KleurringNummer != toupper(raw_KleurringNummer)) %>% 
  distinct(raw_KleurringNummer)
```

are transformed to uppercase:

```{r raw_KleurringNummer_to_upper}
obs_and_acts <- obs_and_acts %>% 
  mutate(raw_KleurringNummer = toupper(raw_KleurringNummer))
```

### Check keys `raw_KleurringNummer` (obs) - `raw_Nummer` (birds)

All values in `raw_KleurringNummer` should be present in `birds$raw_Nummer`. Exceptions:

```{r raw_KleurringNummer_not_in_birds_raw_Nummer}
obs_and_acts %>%
  filter(!raw_KleurringNummer %in% birds$raw_Nummer) %>% 
  distinct(raw_KleurringNummer)
```

We remove them:

```{r remove_obs_and_acts_linked_to_invalid_bird_ring}
raw_KleurringNummer_remove <- obs_and_acts %>%
  filter(!raw_KleurringNummer %in% birds$raw_Nummer) %>% 
  distinct(raw_KleurringNummer) %>%
  pull()
obs_and_acts <- obs_and_acts %>% 
  filter(!raw_KleurringNummer %in% raw_KleurringNummer_remove)
birds <- birds %>%
  filter(!raw_Nummer %in% raw_KleurringNummer_remove)
```

Finally, we search for birds (`raw_Nummer` in  `birds`) which are not present in any observation (`raw_KleurringNummer` in `obs_and_acts`):

```{r birds_not_in_obs}
birds %>% filter(!raw_Nummer %in% obs_and_acts$raw_KleurringNummer)
```

We remove them:

```{r remove_birds_not_in_obs}
birds <- birds %>% 
  filter(raw_Nummer %in% obs_and_acts$raw_KleurringNummer)
```

### Check duplicates in date of applying rings

Detect duplicates in date of applying (the very first) color or metal ring (action code `rngkl` or `rngme`):

```{r duplicates_apply_color_rings}
duplicates_ringing <- obs_and_acts %>%
  filter(raw_AktieCode %in% c("rngkl", "rngme")) %>%
  distinct(raw_KleurringNummer, raw_Datum)  %>%
  group_by(raw_KleurringNummer, raw_Datum) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(obs_and_acts %>%
              filter(raw_AktieCode %in% c("rngkl", "rngme")) %>%
              select(raw_Nummer, raw_KleurringNummer, 
                     raw_Datum, raw_AktieCode)) %>%
  select(-n) %>%
  select(raw_Nummer, everything())
duplicates_ringing
```

### Check age while applying rings

For each bird, observations with same date should refer to same bird age. Exceptions:

```{r check_anomalies_age_ringing}
obs_and_acts %>%
  filter(!is.na(raw_LeeftijdCode)) %>%
  filter(raw_AktieCode %in% c("rngkl", "rngme")) %>%
  distinct(raw_KleurringNummer, raw_Datum, raw_LeeftijdCode)  %>%
  group_by(raw_KleurringNummer, raw_Datum) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(obs_and_acts %>%
              filter(!is.na(raw_LeeftijdCode)) %>%
              select(raw_Nummer, raw_KleurringNummer, raw_Datum, 
                     raw_LeeftijdCode, raw_AktieCode)) %>%
  select(-n) %>%
  select(raw_Nummer, raw_KleurringNummer, raw_Datum, 
         raw_LeeftijdCode, raw_AktieCode)
```

### Check bird sex inconsistencies

All birds should have a sex, which could be one of `M` (_mannetje_), `V` (_vrouwtje_), `O` (_onbekend_):

```{r present_GeslachtCode}
distinct(birds, raw_GeslachtCode)
```

Exceptions:

```{r exceptions_bird_sex}
birds %>%
  filter(is.na(raw_GeslachtCode)) %>%
  distinct(raw_Nummer, raw_NummerNieuw, raw_NummerDesc, raw_GeslachtCode)
```

For all birds a one-to-one relation bird - sex should hold true:

```{r check_one-to-one_bird_sex}
birds %>% 
  distinct(raw_Nummer, raw_GeslachtCode) %>%
  nrow() == nrow(birds)
```

### Handle information about metal ring number

There are columns containing metal ring related information in `birds`:

```{r cols_metal_ring_birds}
names(birds)[which(grepl("raw_Metaalring", names(birds)))]
```

and `obs_and_acts`:

```{r cols_metal_ring_obs_and_acts}
names(obs_and_acts)[which(grepl("raw_Metaalring", names(obs_and_acts)))]
```

This is not only redundant: inconsistencies are detected. Some examples:

```{r inconsisntecy_metal_ring_info_examples}
birds %>%
  filter(raw_Nummer %in% c("AAAA", "AAAR", "AAAT", "AAAW")) %>%
  rename("metal_ring_from_birds"  = "raw_MetaalringNummer") %>%
  left_join(obs_and_acts %>%
              rename("metal_ring_from_obs"  = "raw_MetaalringNummer"),
            by = c("raw_Nummer" =  "raw_KleurringNummer")) %>%
  distinct(raw_Nummer, metal_ring_from_birds, metal_ring_from_obs)
```

Based on experts' knowledge only the metal ring information in `birds` is correct. We remove the columns related to metal ring information from `obs_and_acts` as this information is redundant and not needed for the mapping:

```{r remove_metal_ring_infos_from_obs}
obs_and_acts <-
  obs_and_acts %>%
  select(-starts_with("raw_Metaalring"))
```

### Solve GPS tracker IDs

Some birds have a suspect `raw_Nummer` which is formed by letters `GPS` or `GP` followed by some numbers. These are a kind of GPS IDs which should not be in columns related to color rings:

```{r show_gps_ids}
birds %>%
  filter(str_detect(raw_Nummer, pattern = "(^(GP)\\d+)|(^(GPS)\\d+)")) %>%
  select(raw_Nummer)
```

The right mapping of them is saved in `map_gps`:

```{r preview_map_gps}
map_gps %>% head()
```

We import the color rings in new column `first_Nummer`. For all other rings will be `first_nummer = raw_Nummer` as they are the very first rings. We will also add note `'uvabits_gps_tag'`:

```{r map_to_first_Nummer}
map_gps_add_notes <- map_gps %>%
  mutate(bird_notes = "uvabits_gps_tag.")
birds <- birds %>%
  # create column first_Nummer with right color rings for the GPS or GP** rings
  left_join(map_gps_add_notes, by = c("raw_Nummer" = "gps_id")) %>%
  # set first_Numer equal to raw_Nummer in all other cases
  mutate(first_Nummer = ifelse(is.na(color_ring),
                               raw_Nummer,
                               color_ring)) %>%
  select(raw_Nummer, first_Nummer, everything())
```

Check whether not corrected rings (`GP***` or `GPS***`) are still present:

```{r check_presence_GP_GPS_in_birds}
birds %>%
  filter(str_detect(first_Nummer, pattern = "(^(GP)\\d+)|(^(GPS)\\d+)")) %>%
  select(first_Nummer)
```

Mapping example: the bird with `raw_Nummer` equal to GP130 is mapped as follows:

```{r example_gps_mapping_and_name}
birds %>%
  filter(raw_Nummer == "GP130") %>%
  select(raw_Nummer, first_Nummer, bird_notes)
```

Column `first_Nummer` should contain unique elements. Check for exceptions:

```{r check_uniqueness_first_Nummer}
birds %>%
  group_by(first_Nummer) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(birds) %>%
  select(-n)
```

Exceptions are due to double entry, one with the original color ring, a second one with GPS tracker ID. Such exceptions should be handled separately while mapping `bird_shorthand`.

### Check inconsistencies between users and observers

We have to be sure that all observations have a valid observator reference, i.e. an ID contained in column `raw_Nummer` of `users`.

Observations without an observator reference:

```{r raw_WaarnemerNummer_is_na}
obs_and_acts %>%
  filter(is.na(raw_WaarnemerNummer))
```

All observator references in `obs_and_acts` should be present in `users`:

```{r check_observators_in_users}
observers_ids <- unique(
  obs_and_acts %>%
    filter(!is.na(raw_WaarnemerNummer)) %>%
    distinct(raw_WaarnemerNummer) %>%
    pull())
all(observers_ids %in% users$raw_Nummer)
```

# Map data

We tranform the information extracted from INBO color ring database to SOVON format. The goal is to create three dataframes:

1. `crbirding_users`
2. `crbirding_birds`
3. `crbirding_observations`

## Map user data

We start mapping user data.

### E-mail

```{r e-mail}
users <- users %>%
  mutate(user_email = raw_Email)
```

### First name

```{r first_name}
users <- users %>%
  mutate(user_first_name = raw_Voornaam)
```

### Last name

```{r last_name}
users <- users %>%
  mutate(user_last_name = raw_Familienaam)
```

### Address

```{r address}
users <- users %>%
  mutate(user_address = raw_Adres)
```

### Place

```{r place}
users <- users %>%
  mutate(user_place = raw_Gemeente)
```

### Postal code

```{r postal_code}
users <- users %>%
  mutate(user_postal_code = raw_Postcode)
```

### Country

```{r country}
users <- users %>%
  mutate(user_country = raw_LandCode)
```

### User ID

User identifiers are provided by SOVON. `NA` is given.

```{r userID}
users <- users %>%
  mutate(user_id  = NA)
```

### User reference

We use the unique ID in `raw_Nummer`:

```{r user_ref}
users <- users %>%
  mutate(user_reference  = raw_Nummer)
```

### User language

This field is not present in `users`. However, all accounts have been made by INBO in Dutch:

```{r user_language}
users <- users %>%
  mutate(user_language  = "Dutch")
```

### Create `crbirding_users`

Export the needed columns to `crbirding_users`:

```{r make_crbirding_users}
crbirding_users <- 
  users %>%
  select(-starts_with("raw_"))
```

## Map color ring data

### Bird reference & bird shorthand

The relation color ring (SOVON field `bird_shorthand`) - bird is a multi-to-one relation. By assigning a bird reference as unique integer to a bird we can link any color ring to the bird it belongs to.

```{r bird_reference}
birds <- birds %>%
  mutate(bird_reference = seq_len(nrow(birds))) %>%
  select(bird_reference, first_Nummer, raw_NummerNieuw, 
         raw_NummerDesc, everything())
head(birds)
```

We have to solve the lack of unicity discussed in preprocessing, which ends up in multiple `bird_reference` for the same bird. See here the few issues encountered:

```{r issues_shorthand}
birds %>%
  group_by(first_Nummer) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(birds %>%
              select(bird_reference, 
                     first_Nummer, 
                     raw_NummerNieuw, 
                     raw_NummerDesc, 
                     raw_Nummer),
            by = "first_Nummer") %>%
  select(-n)
```

We assign the lower `bird_reference`:

```{r solve_unicity_issues}
birds <-
  birds %>%
  group_by(first_Nummer) %>%
  mutate(bird_reference = min(bird_reference)) %>%
  ungroup()
```

Check:

```{r check_solution_unicity_shorthand}
birds %>%
  group_by(first_Nummer) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(birds %>%
              select(bird_reference, 
                     first_Nummer, 
                     raw_NummerNieuw, 
                     raw_NummerDesc, 
                     raw_Nummer),
            by = "first_Nummer") %>%
  select(-n)
```

We add column `bird_reference` to `obs_and_acts` as it will be needed to map the dates of ringing. The link is made by the columns `raw_Nummer` (birds data) and  `raw_KleurringNummer` (observation data):

```{r apply_bird_ref_to_obs_acts}
obs_and_acts <- 
  obs_and_acts %>%
  left_join(birds %>% select(raw_Nummer, bird_reference),
            by = c("raw_KleurringNummer" = "raw_Nummer")) %>%
  select(bird_reference, everything())
```

Check whether there are observation without link to any bird:

```{r check_NA_bird_ref}
filter(obs_and_acts, is.na(bird_reference))
```

We will also add column `first_Nummer` from birds to observations as well:

```{r add_first_Nummer_to_obs_and_acts}
obs_and_acts <- 
  obs_and_acts %>%
  left_join(
    birds %>%
      select(bird_reference, first_Nummer, raw_Nummer),
    by = c("bird_reference", 
           "raw_KleurringNummer" = "raw_Nummer")
)
```

The reconstruction of the series of color rings for each bird is made by using columns `first_Nummer`, `raw_NummerNieuw` and the information from spreadsheets of experts. We will collect the needed information in new dataframe, called `crbirding_birds` which will grow up to become the end product of color ring table.

First step is to *gather* `first_Nummer`  and `raw_NummerNieuw` in new column `bird_shorthand`. We do it and we create a new data.frame called `crbirding_birds` which will be the end product related to rings:

```{r backbone_mapping_shorthand}
crbirding_birds <- 
  birds %>%
  select(bird_reference, first_Nummer, 
         raw_NummerNieuw, raw_NummerDesc, bird_notes) %>%
  gather(key = col_nummer,
         value = bird_shorthand,
         first_Nummer, raw_NummerNieuw) %>%
  select(-col_nummer) %>%
  select(bird_reference, bird_shorthand, raw_NummerDesc, bird_notes)
```

We can also assign the color ring version with point (`bird_shorthand_pt`) by matching columns `bird_shorthand` and `raw_NummerDesc` where possible. Note that `bird_shorthand_pt` may contain points, but for old rings it is equal to `bird_shorthand`:

```{r map_shorthand_with_point}
crbirding_birds <-
  crbirding_birds %>%
  mutate(bird_shorthand_pt = ifelse(
    str_remove_all(string = raw_NummerDesc, pattern = "\\.") == bird_shorthand,
    raw_NummerDesc,
    bird_shorthand)
    ) %>%
  distinct() %>%
  select(bird_reference, bird_shorthand, bird_shorthand_pt, bird_notes) %>%
  # to map correctly MKAU (bird_reference: 3716)
  group_by(bird_reference, bird_shorthand) %>%
  filter(bird_shorthand_pt == 
           bird_shorthand_pt[which.max(nchar(bird_shorthand_pt))]) %>%
  ungroup()
```

Example of mapping: the bird with `bird_reference` 14 is associated to the following two rings:

```{r example_mapping_shorthand_birds}
birds %>%
  filter(bird_reference == 14) %>%
  select(bird_reference, first_Nummer, raw_NummerNieuw, raw_NummerDesc)
```

and will be mapped as follows:

```{r example_mapping_shorthand}
crbirding_birds %>%
  filter(bird_reference == 14) %>%
  select(bird_reference, bird_shorthand, bird_shorthand_pt)
```

Another example:

```{r example_bird_ref_3716}
crbirding_birds %>%
  filter(bird_reference == 3716) %>%
  select(bird_reference, bird_shorthand, bird_shorthand_pt)
```

### Bird ringing date

The date of ringing (to be mapped as `bird_date_begin`) can be found in data frame `obs_and_acts` in the column `raw_Datum` relatively to code action `rngkl` (code action of applying colour ring).

```{r example_rngkl_actions}
obs_and_acts %>% 
  filter(raw_AktieCode == "rngkl") %>%
  select(bird_reference, first_Nummer, 
         raw_Datum, raw_AktieCode) %>%
  head(n = 10)
```

We can add automatically the date of the very first ring for each bird (`bird_reference`), based on date of action code `rngkl`. In fact, there should be just one observation with action `rngkl` for each bird (`bird_reference`). Exceptions:

```{r exception_rule_one_bird_one_rngkl}
exceptions_one_bird_one_rngkl <- obs_and_acts %>%
  filter(raw_AktieCode == "rngkl") %>%
  select(bird_reference, raw_Datum) %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(obs_and_acts %>%
              select(bird_reference, raw_Datum, raw_AktieCode),
            by = "bird_reference") %>%
  filter(raw_AktieCode == "rngkl") %>%
  left_join(crbirding_birds %>%
              select(bird_reference, bird_shorthand),
            by = "bird_reference") %>%
  distinct() %>%
  select(-n)
exceptions_one_bird_one_rngkl
```

We will manage the very few exceptions later.

Assign the date of first ringing:

```{r get_first_ring_date}
crbirding_birds <- 
  obs_and_acts %>%
  filter(raw_AktieCode == "rngkl") %>%
  select(bird_reference, first_Nummer, raw_Datum) %>%
  group_by(bird_reference, first_Nummer) %>%
  summarize(bird_date_begin = min(raw_Datum)) %>%
  right_join(crbirding_birds,
             by = c("bird_reference", 
                    "first_Nummer" = "bird_shorthand")) %>%
  rename(bird_shorthand = first_Nummer) %>%
  select(bird_reference, bird_shorthand, bird_shorthand_pt, 
         bird_date_begin, bird_notes)
```

Some examples (birds with `bird_reference` 1, 14 and 4543):

```{r examples_date_ringing}
crbirding_birds %>%
  filter(bird_reference %in% c(1, 14, 4543)) %>%
  arrange(bird_reference, bird_date_begin)
```

Check date of birds in exceptions:

```{r check_date_of_exceptions_double_rngkl}
crbirding_birds %>%
  filter(bird_shorthand %in% c("MKAU", "CZOZ"))
```

And in case solve:

```{r solve_date_of_exceptions_double_rngkl, evaluate = FALSE}
crbirding_birds <- 
  crbirding_birds %>% 
  mutate(bird_date_begin = case_when(
    bird_shorthand == "MKAU" ~ bird_date_begin,
    bird_shorthand ==  "CZOZ" ~ as.POSIXct(NA_character_),
    TRUE ~ bird_date_begin
    )
  )
```

Summary of rings with date (`bird_date_begin`):

```{r summary_n_rings_with_date}
crbirding_birds %>%
  mutate(date_is_present = !is.na(bird_date_begin)) %>%
  group_by(date_is_present) %>%
  count()
```

The rings without date are the last rings of birds ringed more than once. 

```{r summary_bird_with_more_rings}
birds_multiple_rings <- 
  crbirding_birds %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n > 1)
birds_multiple_rings
```

Not only, the structure of our database limits to 2 the maximum number of rings linked to the same bird, as two are the columns containing such information (`raw_Nummer` and `raw_NummerNieuw`):

```{r check_max_rings_is_2}
birds_two_rings <- 
  crbirding_birds %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n == 2)
all(birds_two_rings$bird_reference == birds_multiple_rings$bird_reference)
```

Experts explained us that the very first and the very last ring are mapped in `birds`. By consulting the spreadsheet of the experts we will later fill the gap by adding _intermediate rings_.

We can try to retrieve the date of applying last ring based on observations/actions with code `vang` (caught at the nest) or `vangl` (caught otherwise). Birds ringed twice and linked to one `vang`/`vangl` action:

```{r birds_two_rings_one_vang}
bird_one_vang <- obs_and_acts %>%
  filter(bird_reference %in% birds_multiple_rings$bird_reference,
         raw_AktieCode %in% c("vang", "vangl")) %>%
  select(bird_reference, raw_Datum) %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n == 1) %>%
  left_join(obs_and_acts %>%
              select(bird_reference, raw_Datum, raw_AktieCode),
            by = "bird_reference") %>%
  filter(raw_AktieCode %in% c("vang", "vangl")) %>%
  rename("bird_date_begin" = "raw_Datum") %>%
  arrange(bird_reference)
bird_one_vang
```

As you cannot change a ring without catching him, we can use these unique dates as date of ringing, `bird_date_begin`:

```{r set_date_ringing_bird_catched_once}
bird_one_vang_with_date <- 
  crbirding_birds %>%
  filter(bird_reference %in% bird_one_vang$bird_reference,
         bird_reference %in% birds_two_rings$bird_reference) %>%
  group_by(bird_reference) %>%
  filter(is.na(bird_date_begin)) %>%
  select(-bird_date_begin) %>%
  left_join(bird_one_vang, 
            by = c("bird_reference")
  ) %>%
  select(-c(n, raw_AktieCode))
crbirding_birds <- 
  crbirding_birds %>%
  anti_join(bird_one_vang_with_date,
            by = c("bird_reference", "bird_shorthand")) %>%
  bind_rows(bird_one_vang_with_date) %>%
  arrange(bird_reference)
```

Examples of date mapping (birds with `bird_reference` 11 and 14):

```{r examples_assign_date_second_ring_by_vang_date}
crbirding_birds %>%
  filter(bird_reference %in% c(11, 14))
```

Some birds have been catched twice:

```{r birds_two_rings_two_vang}
bird_two_vang <- obs_and_acts %>%
  filter(bird_reference %in% birds_two_rings$bird_reference,
         raw_AktieCode %in% c("vang", "vangl")) %>%
  select(bird_reference, raw_Datum) %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n == 2) %>%
  left_join(obs_and_acts %>%
              select(bird_reference, raw_Datum, raw_AktieCode),
            by = "bird_reference") %>%
  filter(raw_AktieCode  %in% c("vang", "vangl")) %>%
  rename("bird_date_begin" = "raw_Datum") %>%
  arrange(bird_reference)
bird_two_vang
```

In the expert spreadsheet, we can see that the ring is **typically** applied during the last `vang`/`vangl` action. We will assign the most recent date by default as first step, correcting the exceptions later:

```{r get_second_ring_via_vang}
bird_two_vang_get_date <- 
  crbirding_birds %>%
  filter(bird_reference %in% bird_two_vang$bird_reference,
         bird_reference %in% birds_two_rings$bird_reference) %>%
  group_by(bird_reference) %>%
  filter(is.na(bird_date_begin)) %>%
  select(-bird_date_begin) %>%
  left_join(bird_two_vang %>%
              group_by(bird_reference) %>%
              filter(bird_date_begin == max(bird_date_begin)), 
            by = c("bird_reference")
  ) %>%
  select(-c(n, raw_AktieCode))
bird_two_vang_get_date
```

By consulting the spreadsheet we can find that the following rings have been applied during the earliest `vang`/`vangl` action, so they are exceptions:

```{r correct_zexceptions_bird_date_begin_earliest_date}
bird_shorthand_exceptions <- c("NGAP", "NGAX", "GVAR", 
                               "KPAZ", "KAAK", "KAAN")
bird_exceptions <- 
  crbirding_birds %>%
  filter(bird_shorthand %in% bird_shorthand_exceptions)
```

We set `bird_date_begin` equal to the date of the earliest `vang`/`vangl` action:

```{r}
bird_two_vang_get_date <- 
  bird_two_vang %>%
  filter(bird_reference %in% bird_exceptions$bird_reference) %>%
  group_by(bird_reference) %>%
  filter(bird_date_begin == min(bird_date_begin)) %>%
  left_join(bird_exceptions %>%
              select(-bird_date_begin), by = "bird_reference") %>%
  select(-c(n, raw_AktieCode)) %>%
  select(names(bird_two_vang_get_date)) %>%
  bind_rows(bird_two_vang_get_date %>%
              filter(!bird_shorthand %in% bird_exceptions$bird_shorthand))
```

Add this information to `crbirding_birds`:

```{r}
crbirding_birds <- 
  crbirding_birds %>%
  anti_join(bird_two_vang_get_date,
            by = c("bird_reference", 
                   "bird_shorthand", 
                   "bird_shorthand_pt")) %>%
  bind_rows(bird_two_vang_get_date) %>%
  arrange(bird_reference)
crbirding_birds
```

Rings still without date:

```{r summary_date_begin_after}
crbirding_birds %>%
  mutate(date_is_present = !is.na(bird_date_begin)) %>%
  filter(!date_is_present)
```

Based on spreadsheet and observation data, we found that the dates of applying the rings FAAG and YCAF are linked to actions `ziek`, i.e. the ring has been applied while taking care of the birds. 

The date of applying FAAG:

```{r get_info_FAAG}
info_faag <- 
  crbirding_birds %>%
  filter(bird_shorthand == "FAAG") %>%
  mutate(bird_date_begin = 
           obs_and_acts %>%
           filter(bird_reference == 176, 
                  first_Nummer == "ALAU",
                  raw_AktieCode == "ziek") %>%
           # get the second "ziek" action (2006-07-07)
           filter(raw_Datum == max(raw_Datum)) %>%
           pull(raw_Datum))
info_faag
```

The date of applying YCAF:

```{r get_info_YCAF}
info_ycaf <- 
  crbirding_birds %>%
  filter(bird_shorthand == "YCAF") %>%
  mutate(bird_date_begin = 
           obs_and_acts %>%
           filter(bird_reference == 3293, 
                  first_Nummer == "PLAB",
                  raw_AktieCode == "ziek") %>%
           # there is juwt one "ziek" action
           pull(raw_Datum))
info_ycaf
```

The date of applying FHOV:

```{r get_info_FHOV}
info_fhov <- 
  crbirding_birds %>%
  filter(bird_shorthand == "FHOV") %>%
  mutate(bird_date_begin = 
           obs_and_acts %>%
           filter(bird_reference == 6359, 
                  first_Nummer == "SUAV",
                  raw_AktieCode == "ziek") %>%
           # there is juwt one "ziek" action
           pull(raw_Datum))
info_fhov
```

Add this dates to `crbirding_birds`:

```{r add_to_crbirding_birds}
crbirding_birds <- 
  bind_rows(crbirding_birds %>% 
              filter(!bird_shorthand %in% c("FAAG", "YCAF", "FHOV")),
            info_faag,
            info_ycaf,
            info_fhov) %>%
  arrange(bird_reference, bird_date_begin)
```

Based on spreadsheet of experts, we have to add new rings, as some birds have been ringed more than twice, so the information contained in `first_Nummer` and `raw_NummerNieuw` is not complete, as they map the very first and the very last ring. The intermediate rings are the following:

```{r intermediate_rings_df}
intermediate_rings <- tibble(
    first_Nummer = c(
      "E633",
      "DJAB",
      "BWAD",
      "KLAT",
      "PR3",
      "TY2",
      "RTO",
      "E099",
      "MKAU",
      "ASAH",
      "ADAF"
      ),
    intermediate_ring = c(
      "BUAG",
      "TRAP",
      "GTAS",
      "UPAB",
      "DHAZ",
      "GMAJ",
      "DGAH",
      "GHAT",
      "ZVAU",
      "HWAX",
      "DJAU"
      ),
    bird_date_begin = as.POSIXct(c(
      "2000-05-08", # BUAG
      "2012-05-25", # TRAP
      "2007-06-01", # GTAS
      "2012-05-29", # UPAB
      "2006-05-26", # DHAZ 
      "2007-06-06", # GMAJ
      "2006-05-17", # DGAH
      "2007-06-18", # GHAT
      "2014-05-23", # ZVAU
      "2009-05-18", # HWAX
      "2006-05-26" # DJAU
      ), tz = "UTC")
)
intermediate_rings
```

Retrieve `bird_reference`:

```{r retrieve_bird_ref_intermediate_rings}
intermediate_rings <- 
  intermediate_rings %>%
  left_join(crbirding_birds %>%
              select(bird_reference, bird_shorthand, bird_notes),
            by = c("first_Nummer" = "bird_shorthand")) %>%
  select(bird_reference, first_Nummer, intermediate_ring, 
         bird_date_begin, bird_notes)
intermediate_rings
```

Check whether all dates are linked to a valid observation and check whether they are linked to `vang/vangl` actions:

```{r get_actins_intermediate_rings}
intermediate_rings %>%
  left_join(obs_and_acts %>%
              select(bird_reference, 
                     raw_Datum, 
                     raw_AktieCode, 
                     first_Nummer), 
            by = c("bird_reference", 
                   "bird_date_begin" = "raw_Datum",
                   "first_Nummer")) %>%
  filter(raw_AktieCode %in% c("vang", "vangl"))
```

We can then add the intermediate rings to  `crbirding_birds`:

```{r add_intermediate_ring}
crbirding_birds <- 
  crbirding_birds %>%
  bind_rows(intermediate_rings %>%
              select(-first_Nummer) %>%
              rename(bird_shorthand = intermediate_ring) %>%
              mutate(bird_shorthand_pt = bird_shorthand)) %>%
  arrange(bird_reference, bird_date_begin)
```

Example of a triplet ring sequence:

```{r}
crbirding_birds %>% filter(bird_reference == 7591)
```

At this point all rings should have a `bird_date_begin`:

```{r final_check_all_rings_with_date}
crbirding_birds %>%
  filter(!is.na(bird_date_begin)) %>%
  nrow() == nrow(crbirding_birds)
```

### Bird ringing end date

For birds ringed more than once, we can assign an end date for the changed rings. This information will be stored in field `bird_date_end`. The end date is equal to the date of applying the new ring

```{r assign_bird_date_end}
assign_end_date <- function(data) {
  if (nrow(data) > 1) {
    return(c(as_date(data$bird_date_begin[2:nrow(data)], tz = "UTC"), 
             as_date(NA, tz = "UTC")))
  } else {
    return(as_date(NA, tz = "UTC"))
  }
}

crbirding_birds <- 
  crbirding_birds %>%
  arrange(bird_reference, bird_date_begin) %>%
  group_by(bird_reference) %>%
  nest() %>%
  mutate(bird_date_end = map(data, assign_end_date)) %>%
  unnest() %>%
  select(bird_reference, bird_shorthand, bird_shorthand_pt, 
         bird_date_begin, bird_date_end, bird_notes)
```

Some examples:

```{r examples_bird_date_end}
crbirding_birds %>%
  filter(!is.na(bird_date_end)) %>%
  head(n = 10)
```

### Ring number

The _ring number_, or _metal ring number_, can be found in field `raw_MetaalringNummer`. Ring number should consist of up to ten **alphanumeric** characters. At page 7 of the [EURING Exchange Code 2000+ document](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) we read the following instructions:

> Where the ring number consist of fewer than ten numbers/letters, the number is padded with dots. These dots are always inserted to the left of the rightmost
row of numbers.

Ring number anomalies:

```{r anomalies_ring_number}
birds %>%
  filter(!str_detect(raw_MetaalringNummer, pattern = regex("\\d+"))) %>%
  group_by(raw_MetaalringNummer) %>%
  count()
```

Value `none` means _not metal-ringed bird_, while `onbekend` means that information is no more available. SOVON chooses to leave column `bird_ring_number` empty for both cases. However, we will add a note to `bird_notes` to still maintain a reference to this slight difference. Value `?` is equivalent to `onbekend` while `H????????` and `Lxxxxxx` are mapped as `H--------` and `L------` respectively.

```{r solve_anomalies_bird_ring_number}
birds <- 
  birds %>%
  mutate(bird_ring_number = recode(
    raw_MetaalringNummer,
    "?" = NA_character_,
    "onbekend" = NA_character_,
    "none" = NA_character_,
    "H????????" = "H--------",
    "Lxxxxxx" = "L------"
    )
)
```

We separate the birds with special mapping values by all the others:

```{r hold_special_values}
special_values <- c("H--------", "L------")
birds_special_values <- 
  birds %>%
  filter(bird_ring_number %in% special_values)
birds_others <- 
  birds %>%
  filter(!bird_ring_number %in% special_values)
```

Some ring numbers contain asterisks:

```{r show_asterisks}
birds_others %>% 
  filter(str_detect(bird_ring_number, "\\*")) %>%
  select(bird_ring_number)
```

We remove the asteriks:

```{r remove_*_in_ring_number}
birds_others <- 
  birds_others %>%
  mutate(bird_ring_number = gsub("\\*", "", bird_ring_number))
```

Add points `.` where needed:

```{r add_._in_ring_number}
birds_others <- 
  birds_others %>%
  mutate(reversed = stri_reverse(str = bird_ring_number))
idx <- as.data.frame(str_locate(birds_others$reversed, pattern = regex("\\d+")))$end
idx <- replace_na(idx, 0)
birds_others <- birds_others %>%
  mutate(bird_ring_number = stri_reverse(str_c(
    str_sub(reversed, 
          end = idx),
  map_chr(nchar(reversed), function(x) {
    ifelse(x < 10,
           str_c(rep(".", 10 - x), collapse = ""),
           "")
  }),
  str_sub(reversed, 
          start = idx + 1)))) %>%
  select(-reversed)
```

Merge the two data frames together:

```{r merge_back_together}
birds <- bind_rows(birds_others, birds_special_values)
```

Effects of mapping (part of it):

```{r show_effect_mapping_example}
birds %>% 
  distinct(raw_MetaalringNummer, bird_ring_number) %>%
  head(n = 200)
```

We have to be sure that no multiple metal rings are assigned to the same bird (`bird_reference`). Exceptions:

```{r multiple_metal_rings_bird_reference}
birds %>%
  select(bird_reference, bird_ring_number) %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(birds) %>%
  select(-n)
```

Thanks to observations with action `rngme` we can retrieve the date of applying the metal ring.

```{r example_apply_metalring}
obs_and_acts  %>%  
  filter(raw_AktieCode == "rngme") %>%
  select(bird_reference, raw_KleurringNummer, raw_Datum, raw_AktieCode) %>%
  head(n = 20)
```

Similarly to color rings and action `rngkl`, there should be one and only one action `rngme` for each bird. Exceptions:

```{r exceptions_one_bird_one_rngme}
obs_and_acts %>%
  filter(raw_AktieCode == "rngme") %>%
  select(bird_reference, raw_KleurringNummer, raw_Datum, raw_AktieCode) %>%
  group_by(bird_reference) %>%
  count() %>%
  filter(n > 1) %>%
  left_join(obs_and_acts %>%
              filter(raw_AktieCode == "rngme") %>%
              select(bird_reference, raw_KleurringNummer, 
                     raw_Datum, raw_AktieCode),
            by = "bird_reference") %>%
  select(-n)
```

Some metal rings are not linked to an observation with action `rngme`, so we have no date for them:

```{r check_apply_rngme}
no_rngme <- 
  birds %>%
  filter(!is.na(bird_ring_number)) %>%
  select(bird_reference, bird_ring_number) %>%
  left_join(obs_and_acts %>%
              filter(raw_AktieCode == "rngme") %>%
              select(bird_reference, raw_Datum),
            by = "bird_reference") %>%
  filter(is.na(raw_Datum)) %>%
  left_join(birds %>%
              select(bird_reference, 
                     raw_Nummer,
                     raw_NummerNieuw, 
                     raw_NummerDesc,
                     raw_MetaalringNummer),
            by = "bird_reference") %>%
  select(bird_reference, raw_Nummer, raw_MetaalringNummer) %>%
  arrange(raw_Nummer)
no_rngme
```

Experts are sure that these metal rings have been applied while applying the color ring. So, we can use the date related to action `rngkl`:

```{r add_date_to_metal_ring_number_exceptions}
no_rngme_date <- 
  no_rngme %>%
  left_join(obs_and_acts %>%
              filter(raw_AktieCode == "rngkl") %>%
              select(bird_reference, raw_Datum),
            by = "bird_reference")
no_rngme_date
```

Be sure this solution provides a date for all metal rings:

```{r check_date_for_all_metal_rings}
no_rngme_date %>%
  filter(!is.na(raw_Datum)) %>%
  nrow() == nrow(no_rngme_date)
```

Check also that all `bird_reference` are linked to one and only one `rngme` action:

```{r check_no_ambiguity_rngkl}
nrow(no_rngme_date) == nrow(no_rngme)
```

If `TRUE`, then we can proceed to add the field `bird_ring_number` to `crbirding_birds` based on:

1. `bird_reference` 
2. Date of applying the metal rings
3. Date of applying the color rings 
3. End date of the color rings (if exists)

This mapping takes into account the fact that some metal rings could have been applied while applying a second color ring or even after.

We first retrieve dates of applying metal rings:

```{r retrieve_date_rngme}
obs_and_acts_rngme <- obs_and_acts %>%
  filter(raw_AktieCode == "rngme") %>%
  select(bird_reference, raw_Datum) %>%
  bind_rows(no_rngme_date %>%
              select(bird_reference, raw_Datum))
```

in order to fill the field `bird_ring_number`:

```{r}
crbirding_birds <- 
  obs_and_acts_rngme %>%
  left_join(birds %>%
              select(bird_reference, bird_ring_number),
            by = "bird_reference") %>%
  rename(bird_date_begin_metalring = raw_Datum) %>%
  right_join(crbirding_birds, by = "bird_reference") %>%
  mutate(bird_ring_number = 
           if_else(is.na(bird_date_end) | 
                     (bird_date_begin_metalring < bird_date_end),
                   bird_ring_number,
                   NA_character_)) %>%
  # remove help column after use
  select(-bird_date_begin_metalring) %>%
  # provide order columns
  select(bird_reference, bird_shorthand, bird_shorthand_pt,
         bird_ring_number, bird_date_begin, bird_date_end,
         bird_notes)
```

Two examples:

```{r}
crbirding_birds %>%
  filter(bird_reference %in% c(8999, 9028))
```

Add note about absence of metal ring based on original values in column `raw_MetaaalringNummer` and the fields `bird_reference`, `bird_ring_number`:

```{r add_note_absence_metal_ring}
crbirding_birds <-
  crbirding_birds %>%
  left_join(birds %>%
              select(bird_reference, 
                     raw_MetaalringNummer),
            by = c("bird_reference")) %>% 
  mutate(bird_notes = case_when(
     is.na(bird_ring_number) & 
       (raw_MetaalringNummer %in% c("?", "onbekend") | 
         is.na(raw_MetaalringNummer)) ~ 
       ifelse(is.na(bird_notes),
              "bird_ring_number not available.",
              str_c(bird_notes, "bird_ring_number not available.", sep = " ")),
     is.na(bird_ring_number) & raw_MetaalringNummer == "none" ~ 
      ifelse(is.na(bird_notes),
             "bird_ring_number not present.",
             str_c(bird_notes, "bird_ring_number not present.", sep = " ")),
     TRUE ~ bird_notes)) %>%
  select(-raw_MetaalringNummer)
```

This is how `bird_notes` has been updated, limited to rows where changes could happen:

```{r updated_bird_notes}
crbirding_birds %>% 
       filter(is.na(bird_ring_number)) %>%
       select(bird_reference) %>%
       left_join(crbirding_birds,
                 by = "bird_reference") %>%
  select(-bird_shorthand_pt)
```

Notice how nothing has been added in `bird_notes` for birds getting a metal ring while changing the color ring.

Based on code actions `klgev` (only colour ring found) and `meweg` (metal ring missing) we can retrieve the end date of some metal rings:

```{r example_klgev}
obs_and_acts  %>%  
  filter(raw_AktieCode == "klgev") %>%
  select(bird_reference, raw_KleurringNummer, raw_Datum, raw_AktieCode) %>%
  head(n = 20) %>%
  arrange(bird_reference, raw_Datum)
```

```{r example_meweg}
obs_and_acts  %>%  
  filter(raw_AktieCode == "meweg") %>%
  select(bird_reference, raw_KleurringNummer, raw_Datum, raw_AktieCode) %>%
  head(n = 20) %>%
  arrange(bird_reference, raw_Datum)
``` 

Waiting answer from INBO experts about the mapping of these actions: see [issue 34](https://github.com/inbo/sovon/issues/34).

### Euring

Present values:

```{r euring_in_db}
birds %>% distinct(raw_EuringCode)
```

We assign the euring codes by applying the following mapping:

```{r bird_euring}
crbirding_birds <-
  crbirding_birds %>%
  left_join(birds %>% 
              distinct(bird_reference, raw_EuringCode),
            by = "bird_reference") %>%
  mutate(bird_euring = recode(raw_EuringCode,
                              "5920" = "05920",
                              "5910" = "05910",
                              "5926" = "05926",
                              "5922" = "05922",
                              "zz" = NA_character_,
                              "zzz" = NA_character_)) %>%
  ungroup()
```

Effects of mapping:

```{r comparison_euring}
crbirding_birds %>% 
  distinct(raw_EuringCode, bird_euring)
```

We remove the help column `raw_EuringCode` from `crbirding_birds`:

```{r remove_raw_euringCode}
crbirding_birds <- 
  crbirding_birds %>%
  select(-raw_EuringCode)
```

### Scheme

Actual values:

```{r scheme}
birds %>% distinct(raw_MetaalringLandCode)
```

We apply the following mapping:

```{r scheme_mapping}
crbirding_birds <- 
  crbirding_birds %>%
  left_join(birds %>% 
              select(bird_reference, raw_MetaalringLandCode),
            by = "bird_reference") %>%
  mutate(bird_scheme = recode(raw_MetaalringLandCode,
                              "BE " = "BLB",
                              "FR" = "FRP",
                              "NL" = "NLA",
                              "PT" = "POL",
                              "UK" = "GBT"))
```

Effects of mapping:

```{r comparison_scheme}
crbirding_birds %>% distinct(raw_MetaalringLandCode, bird_scheme)
```

We remove the help column `raw_MetaalringLandCode` from `crbirding_birds`:

```{r remove_raw_MetaalringLandCode}
crbirding_birds <- 
  crbirding_birds %>%
  select(-raw_MetaalringLandCode)
```

### Bird sex

Bird sex is translated to English. Letter `M` (Dutch word _mannetje_) will not change so no need to convert it:

```{r sex}
crbirding_birds <- 
  crbirding_birds %>%
  left_join(birds %>% 
              select(bird_reference, raw_GeslachtCode),
            by = "bird_reference") %>%
  mutate(bird_sex = recode(raw_GeslachtCode,
    "V" = "F", # V(rouwtje) ->F(emale)
    "O" = "U" # O(nbekend) ->  U(nknown)
    ))
```

Effects of mapping:

```{r comparison_sex}
crbirding_birds %>% distinct(raw_GeslachtCode, bird_sex)
```

We remove the help column `raw_GeslachtCode` from `crbirding_birds`:

```{r remove_raw_GeslachtCode}
crbirding_birds <- 
  crbirding_birds %>%
  select(-raw_GeslachtCode)
```

### Bird age ringing

Mapping the age while applying color ring from INBO format to SOVON standard (Euring standard: see [online pdf document](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) at page 14). Bird age at the moment of any observation can be found in column `raw_LeeftijdCode` of `obs_and_acts`:

```{r map_age_code}
obs_and_acts <- 
  obs_and_acts %>%
  mutate(bird_age_ringing = recode(raw_LeeftijdCode,
    "PU" = "1",
    "AD" = "A",
    "J1" = "3",
    "I4" = "5",
    "I3" = "7",
    "I2" = "9",
    "I5" = "B",
    .missing = "O"))
```

For mapping the bird age while applying color ring we should not only take into account observations with actions `rngkl` (applying very first color ring) but also actions `vang` and `vangl` as applying a new color ring is linked to one of this two actions (see section [bird shorthand](#bird-shorthand)). The age can be added to `crbirding_birds` by matching `bird_date_begin` of `crbirding_birds` and `raw_Datum` of `obs_and_acts` for each `bird_reference`. First we check that this strategy ends up with an one-to-one relation:

```{r check_uniqueness_approach}
crbirding_birds %>%
  left_join(obs_and_acts %>%
              distinct(bird_reference, raw_Datum, 
                       bird_age_ringing, raw_AktieCode) %>%
              # remove rows with raw_Datum equal to NA or field observations
              filter(!is.na(bird_age_ringing), 
                     raw_AktieCode %in% c("vang", "vangl", "rngkl")),
            by = c("bird_reference", "bird_date_begin" = "raw_Datum")) %>% 
  nrow() == nrow(crbirding_birds)
```

If TRUE then, we proceed with the mapping:

```{r apply_mapping_bird_age_ringing}
crbirding_birds <- 
  crbirding_birds %>%
  left_join(obs_and_acts %>%
              distinct(bird_reference, raw_Datum, 
                       bird_age_ringing, raw_AktieCode) %>%
              # remove rows with raw_Datum equal to NA or field observations
              filter(!is.na(bird_age_ringing), 
                     raw_AktieCode %in% c("vang", "vangl", "rngkl")),
            by = c("bird_reference", "bird_date_begin" = "raw_Datum")) %>%
  select(-raw_AktieCode)
```

### Bird ID

Bird identifiers will be provided by SOVON. `NA` is given:

```{r birdID}
crbirding_birds <- 
  crbirding_birds %>%
  mutate(bird_id = NA)
```

### Bird BTO

Bird BTO will be provided by SOVON.  `NA` is given:

```{r bird_bto}
crbirding_birds <- 
  crbirding_birds %>%
  mutate(bird_bto = NA)
```

### Bird name

Some birds, the ones with a GPS tracker, have a name saved in [this file](https://raw.githubusercontent.com/inbo/bird-tracking/master/data/interim/individuals.csv). The names are saved in column `individual_remarks`:

```{r read_names}
uvabits_file <- "https://raw.githubusercontent.com/inbo/bird-tracking/master/data/interim/individuals.csv"
uvabits_names <- 
  read_csv(uvabits_file) %>%
  select(individual_id, ring_number, individual_remarks, everything())
head(uvabits_names)
```

We couple the birds to the correspondent names:

```{r coupling_bird_reference_name}
bird_ref_uvabits <- 
  crbirding_birds %>%
  left_join(uvabits_names %>%
              select(colour_ring, individual_remarks), 
            by = c("bird_shorthand" = "colour_ring")) %>%
  rename("bird_name" = "individual_remarks") %>%
  filter(!is.na(bird_name)) %>%
  select(bird_reference, bird_name)
```

And we assign the names by joining on `bird_reference`:

```{r add_names}
crbirding_birds <-
  crbirding_birds %>%
  left_join(bird_ref_uvabits, by = "bird_reference")
```

### Bird birth year

This field will be filled by SOVON. `NA` is given:

```{r bird_birth_year}
crbirding_birds <- 
  crbirding_birds %>%
  mutate(bird_birth_year = NA)
```

## Finalize `crbirding_users`

We have now sufficient information to define the role of any user in `crbirding_users`.

### Add user role to `crbirding_users`

Based on `bird_date_begin` we can retrieve the observations linked to apply rings. These observations can be labelled as `rngkl` (applying very first color ring), but they can sometimes come from `vang` or `vangl` as we have discussed before. We retrieve the ringers by joining on  `raw_WaarnemerNummer` between `crbirding_users` and `obs_and_acts`:

```{r find_ringers}
ringers_number <- 
  crbirding_birds %>%
  select(bird_reference, bird_date_begin) %>%
  left_join(obs_and_acts %>% 
              filter(raw_AktieCode %in% c("rngkl", "rngme", "vang", "vangl")) %>%
              select(bird_reference, raw_Datum, raw_WaarnemerNummer),
            by = c("bird_reference", "bird_date_begin" = "raw_Datum")) %>%
  filter(!is.na(raw_WaarnemerNummer)) %>%
  distinct(raw_WaarnemerNummer) %>%
  pull(raw_WaarnemerNummer)
print(paste("Number of ringers:", 
            length(ringers_number)))
```

We assign a `R` (ringer) to them, `O` (observer) otherwise:

```{r assign_ringer}
crbirding_users <- 
  crbirding_users %>%
  mutate(user_role = ifelse(user_reference %in% ringers_number,
                      "R", "O"))
```

Number of ringers and observers:

```{r summary_n_ringers_observers}
crbirding_users %>%
  group_by(user_role) %>%
  count()
```

## Map observation data

### Observation reference

We add an id number to each observation:

```{r obs_ref}
obs_and_acts <- 
  obs_and_acts %>% 
  mutate(observation_reference = seq_len(nrow(obs_and_acts)))
```

### Observation date

```{r obs_date}
obs_and_acts <- obs_and_acts %>% mutate(observation_date = raw_Datum)
```

### Observation time

No time is found for observations in `obs`. `NA` is given:

```{r obs_time}
obs_and_acts <- obs_and_acts %>% mutate(observation_time = NA)
```

### Observation latitude

Observation latitude should be converted to decimal degrees:

```{r obs_latitude}
obs_and_acts <- obs_and_acts %>% mutate(
  observation_lat = conv_unit(
    str_c(raw_PlaatsBreedtegraadGraden,
          raw_PlaatsBreedtegraadMinuten,
          raw_PlaatsBreedtegraadSeconden, 
          sep = " "), 
    from = 'deg_min_sec', to = 'dec_deg'
))
```

### Observation longitude

Observation longitude should be converted to decimal degrees:

```{r obs_longitude}
obs_and_acts <- obs_and_acts %>% mutate(
  observation_long = conv_unit(
    str_c(raw_PlaatsLengtegraadGraden,
          raw_PlaatsLengtegraadMinuten,
          raw_PlaatsLengtegraadSeconden, 
          sep = " "), 
    from = 'deg_min_sec', to = 'dec_deg'
))
```

### Observation location

Aggregate information about observation location. We follow the following structure: `raw_PlaatsGemeente` [+ `, ` + `raw_PlaatsToponym` [+ `: ` + `raw_PlaatsToponymDetail`]]:

```{r}
obs_and_acts <- obs_and_acts %>% mutate(
  observation_location = ifelse(!is.na(raw_PlaatsToponym),
                                str_c(raw_PlaatsGemeente,
                                      raw_PlaatsToponym, 
                                      sep = ", "),
                                paste(raw_PlaatsGemeente))) %>%
  mutate(observation_location = ifelse(!is.na(raw_PlaatsToponymDetail),
                                       str_c(observation_location,
                                             raw_PlaatsToponymDetail,
                                             sep = ": "),
                                       paste(observation_location)))
```

### Observation status

The need of adding the EURING status information arises by the need of mapping the observations with action code `br` and `vang`:

```{r breed_show}
actions_meaning %>%
  filter(Code %in% c("br", "vang"))
```

The actions `br` and `vang` refer to status `breeding` in EURING system. Based on the scheme at page 16 of [EURING Exchange Code 2000+](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf) document, we define the following mapping:

```{r}
obs_and_acts %>%
  mutate(status = case_when(
    raw_AktieCode %in% c("br", "vang") ~ "N",
    raw_LeeftijdCode == "PU" ~ "-",
    TRUE ~ "U"
  ))
```

### Observation reporter

The field `melder` is equal to `user_first_name` and `user_last_name` in `users`:

```{r obs_melder}
obs_and_acts <- obs_and_acts %>%
  left_join((users %>% select(raw_Nummer, user_first_name, user_last_name)),
            by = c("raw_WaarnemerNummer" = "raw_Nummer")) %>%
  mutate(melder = str_c(user_first_name, user_last_name, sep = " ")) %>%
  select(-c(user_first_name, user_last_name))
```

### Observation reporter reference number

The field `melder_ringersnummer` is equal to `user_reference` and should be left to SOVON. `NA` is given:

```{r melder_ringersnummer}
obs_and_acts <- obs_and_acts %>%
  mutate(melder_ringersnummer = NA)
```

### Observation notes

```{r obs_notes}
obs <- obs %>% mutate(observation_notes = raw_Opmerking)
```

### MRI: metal ring information

Metal ring information table can be found at page 8 of [this pdf document](https://euring.org/files/documents/E2000PLUSExchangeCodev1161.pdf)

A summary of the mapping follows:

1. observations with action code `meweg` should be mapped to 0 (metal ring is not present)

Further mapping steps need the knowledge of an expert.

```{r MRI}

```

# Post-processing

## User data 

Are all required columns present?

```{r check_presence_required_cols_users}
all(cr_users_cols %in% names(crbirding_users))
```

Arrange columns `crbirding_users`:

```{r get_right_order_cols_users}
crbirding_users <-
  crbirding_users %>%
  select(cr_users_cols)
```

Preview data (e-mail, first and last names removed for privacy reasons):

```{r final_preview_users}
crbirding_users %>%
  select(-c(user_email, user_first_name, user_last_name)) %>%
  head(n = 10)
```

Save to CSV:

```{r write_user_data_txt}
write_csv(crbirding_users, path = cr_users_file, na = "")
```

## Bird data

SOVON is interested in color rings versions with a dot in case dots are present, even if the dots are for readibility reasons. At the end of the mapping we can overwrite `bird_shorthand` with `bird_shorthand_pt`:
```{r remove_col_bird_shorthand_pt}
crbirding_birds <- 
  crbirding_birds %>%
  select(-bird_shorthand) %>% 
  rename(bird_shorthand = bird_shorthand_pt)
```

Are all required columns present?

```{r check_presence_required_cols_birds}
all(cr_birds_cols %in% names(crbirding_birds))
```

Arrange columns `crbirding_birds`:

```{r get_right_order_cols_birds}
crbirding_birds <-
  crbirding_birds %>%
  select(cr_birds_cols)
```

Preview data:

```{r final_preview_birds}
crbirding_birds %>% head(n = 10)
```

Save to CSV:

```{r write_bird_data_txt}
write_csv(crbirding_birds, path = cr_birds_file, na = "")
```

## Observation data

Are all required columns present?

```{r check_presence_required_cols_obs}
all(cr_obs_cols %in% names(crbirding_obs))
```

Arrange columns `crbirding_birds`:

```{r get_right_order_cols_obs}
crbirding_obs <-
  crbirding_obs %>%
  select(cr_obs_cols)
```

Preview data:

```{r final_preview_obs}
crbirding_obs %>% head(n = 10)
```

Save to CSV:

```{r write_obs_data_txt}
write_csv(crbirding_obs, path = cr_obs_file, na = "")
```
